---
title: "Wrangling Data with the Tidyverse"
author: "Computer Applications for the Psychological Sciences"
date: "Fall 2025"
output:
  slidy_presentation: default
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, fig.align = "center")
library(tidyverse)
library(janitor)
library(skimr)
library(broom)
pct <- function(x, digits = 1) paste0(round(100*x, digits), "%")
```

# What is Data Wrangling?
* The process of transforming raw, messy data into a clean, organized, and usable format.

* Analogy: Like a chef preparing ingredients before cooking a meal, data wrangling prepares data for analysis.

# The 80/20 Rule of Data Science
* The Rule: Data professionals spend approximately 80% of their time discovering, cleaning, and reorganizing data.

* Only 20% of their time is spent on the "core" tasks of analysis and modeling.

* Why? Data in the wild is almost never in a format ready for analysis. This foundational work is critical for meaningful results.

# The Tidyverse: A Grammar of Data Manipulation
The tidyverse is a collection of R packages (ggplot2, dplyr, tidyr, etc.) that are designed for data science and share a common philosophy and grammar, making data work more efficient and intuitive.

Its most important operator is the pipe (%>%).

# The Pipe Operator (%>%) {.smaller}
The pipe operator allows you to chain functions together, passing the result of one function as the first argument to the next. You can read %>% as "and then".

## Without the pipe (hard to read):
```{r}
arrange(summarise(group_by(storms, year), avg_wind = mean(wind)), desc(avg_wind))
```

# With the pipe (clean and sequential):
```{r}
storms %>%
  group_by(year) %>%
  summarise(avg_wind = mean(wind)) %>%
  arrange(desc(avg_wind))
```

# What is Tidy Data?
Tidy data is a standardized format that makes data easier to model and visualize. It follows three simple rules:

* Every column is a variable.

* Every row is an observation.

* Every cell is a single value.

Raw data is often "untidy." Our goal is to use tidyverse tools to make it tidy.

# A Common First Step: clean_names()

Real-world datasets often have messy column names (First Name, dollar_amount ($)). The clean_names() function from the janitor package is a lifesaver.

It automatically:

* Converts all names to snake_case

* Removes special characters

* Handles spaces and capitalization

```{r}
# Imagine a messy data frame
#messy_data %>% 
#  clean_names() # That's it! Your column names are now clean.
```

# Core Verbs:

## select()
* To subset your data by keeping or removing columns.

## filter()
* To subset your data by keeping rows that meet specific logical conditions.

## mutate()
* To create new columns, often by performing calculations on existing columns.

# select()

Example: Let's select only four columns from the storms dataset.

```{r}
storms %>%
  select(name, year, wind, pressure) %>%
  head(5) # Show first 5 rows
```

# filter()

Example: Let's find all storms in 1995 that were hurricane strength (wind >= 74 knots).
```{r}
storms %>%
  filter(year == 1995, wind >= 74) %>%
  head(5) # Show first 5 rows
```

# mutate()

Example: Let's convert wind speed from knots to kph (1 knot ≈ 1.852 kph).
```{r}
storms %>%
  select(name, wind) %>%
  mutate(wind_kph = wind * 1.852) %>%
  head(5) # Show first 5 rows
```

# Your Turn! (Practice Problem)

## Question: Using the storms dataset, can you find the name, year, and pressure for all storms that occurred in the 21st century (year >= 2000) and had a pressure less than 950mb?

**Hint: You will need to use filter() and select().**

# The Power Couple: group_by() + summarise()
**group_by()**

* Groups the data so that subsequent verbs operate on a per-group basis.

**summarise()**

* Collapses each group into a single-row summary.

Example: Let's find the maximum wind speed for each year.
```{r}
storms %>%
  group_by(year) %>%
  summarise(max_wind = max(wind)) %>%
  head(5) # Show first 5 years
```

# Reshaping Data: From Wide to Tidy
Problem: Sometimes variables are stored as column names. This is "untidy."
Solution: pivot_longer() collapses multiple columns into key-value pairs.

Before: The who dataset is wide
```{r}
#Select a few columns to show the "wide" structure
who %>% 
  select(country, new_sp_m014, new_sp_m1524) %>% 
  head(3)
```

# Applying pivot_longer()
We tell pivot_longer which columns to gather (cols), where to put the old column names (names_to), and where to put the cell values (values_to).

After: The data is "longer" and tidy
```{r}
# The code to make the 'who' dataset tidy
tidy_who <- who %>%
  pivot_longer(
    cols = new_sp_m014:newrel_f65,
    names_to = "key",
    values_to = "count"
  )

#Take a peek at our new variables
tidy_who %>% 
  select(country, key, count) %>% 
  head(5)
```


# Case Study: Finding the Windiest Year
Question: Which year had the highest average maximum wind speed?

**Strategy:**

1. Group the data by year.

2. Summarise each year to find its average maximum wind.

3. Arrange the results and take the top one.

**The single, chained command gives us our answer**
```{r}
storms %>%
  group_by(year) %>%
  summarise(avg_max_wind = mean(wind, na.rm = TRUE)) %>%
  slice_max(order_by = avg_max_wind, n = 1)
```
