---
title: "Quiz: Data Wrangling with the Tidyverse (with Solutions)"
author: ""
date: ""
output:
  pdf_document:
    toc: false
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
```

# Shared mini-datasets (used across questions)

```{r data}
sales <- tibble::tribble(
  ~order_id, ~date,        ~store, ~category,   ~units, ~revenue,
  "A001",    "2025-01-01", "WBG",  "Coffee",     2,     8.50,
  "A002",    "2025-01-01", "WBG",  "coffee ",    1,     4.25,
  "A003",    "2025-01-02", "WBG",  "Tea",        NA,    3.75,
  "A004",    "2025-01-02", "RVA",  "COFFEE",     3,     NA,
  "A005",    "2025-01-03", "RVA",  "Tea",        2,     7.50,
  "A006",    "2025-01-03", "WBG",  "Snacks",     5,     12.00
)

weather <- tibble::tribble(
  ~date,        ~temp_f, ~precip_in,
  "2025-01-01",  42,     0.00,
  "2025-01-02",  39,     0.20,
  "2025-01-03",  45,     NA
)

wide_temps <- tibble::tribble(
  ~day, ~Temp_Month_1, ~Temp_Month_2,
   1,        42,           NA,
   2,        39,           NA,
   3,        45,           NA,
   1,        NA,           50,
   2,        NA,           48
)
```

\newpage

# Questions

## 1) Import + types (very brief code)

The `sales$date` column is currently character. Write **one line** of code to convert it to a `Date` column in the `sales` tibble.

```{r q1, eval=FALSE}
# YOUR CODE
```

---

## 2) Count categories (very brief code)

Write **one line** of code to produce a frequency table of `category` values in `sales`, sorted from most common to least common.

```{r q2, eval=FALSE}
# YOUR CODE
```

---

## 3) String cleaning (very brief code)

Write a `mutate()` line that creates `category_clean` by:
- lowercasing,
- trimming whitespace,
- and converting to title case (e.g., `"coffee"` → `"Coffee"`).

```{r q3, eval=FALSE}
# YOUR CODE
```

---

## 4) Missingness diagnostics (very brief code)

Write code to compute **percent missing** for `revenue` in `sales` (a single number).

```{r q4, eval=FALSE}
# YOUR CODE
```

---

## 5) Filtering missing values (brief code)

Write code to remove rows where `units` is missing **or** `revenue` is missing.

```{r q5, eval=FALSE}
# YOUR CODE
```

---

## 6) Summarise correctly (identify and fix an error)

This code errors or returns the wrong result:

```r
sales %>%
  summarise(avg_rev = mean(revenue))
```

Write the corrected version so it returns the average revenue **ignoring missing values**.

```{r q6, eval=FALSE}
# YOUR CODE
```

---

## 7) Grouped summaries (brief code)

Write code to compute `avg_revenue` by `store` (one row per store), ignoring missing revenue.

```{r q7, eval=FALSE}
# YOUR CODE
```

---

## 8) `na.rm` misconception (short written response)

A student writes:

```r
sales %>%
  summarise(
    total_units = sum(units, na.rm = TRUE),
    avg_revenue = mean(revenue, na.rm = TRUE)
  )
```

In **1–2 sentences**, describe **one important limitation** of relying only on `na.rm = TRUE` for missing data handling.

**Your answer:**

---

## 9) Join keys and type mismatch (identify + fix)

A student tries:

```r
sales %>%
  left_join(weather, by = "date")
```

but gets lots of missing `temp_f` after the join.

Give **one likely reason** based on this dataset and show **one line** that fixes it.

```{r q9, eval=FALSE}
# YOUR CODE + BRIEF NOTE
```

---

## 10) Validate a join with `anti_join()` (brief code)

Write code to identify dates that appear in `sales` but do **not** appear in `weather`.

```{r q10, eval=FALSE}
# YOUR CODE
```

---

## 11) Join explosion detection (brief code)

Write code that checks whether `weather` has duplicate join keys on `date`. If duplicates exist, the output should show them.

```{r q11, eval=FALSE}
# YOUR CODE
```

---

## 12) Reshape wide → long (brief code)

Using `wide_temps`, write code to create a long table with columns:
- `day`
- `month` (values like `1`, `2`)
- `temp`

```{r q12, eval=FALSE}
# YOUR CODE
```

---

## 13) Remove pivot artifacts (brief code)

After reshaping `wide_temps` to long, many rows will have `temp = NA`. Write **one line** of code to remove those rows.

```{r q13, eval=FALSE}
# YOUR CODE
```

---

## 14) `case_when()` recoding (brief code)

Write code to create `store_type` in `sales` where:
- `"WBG"` → `"Williamsburg"`
- `"RVA"` → `"Richmond"`
- otherwise → `"Other"`

```{r q14, eval=FALSE}
# YOUR CODE
```

---

## 15) Common grouping mistake (spot and correct)

A student expects **one row total**, but gets two:

```r
sales %>%
  group_by(store) %>%
  summarise(n = n())
```

They later run:

```r
sales %>%
  group_by(store) %>%
  summarise(n = n()) %>%
  summarise(total = sum(n))
```

1. In **1–2 sentences**, explain why they got two rows in the first step.  
2. Provide a **single pipeline** that returns **one row** with:
   - `n_total` (total rows),
   - `n_wbg` (rows where store == "WBG"),
   - `n_rva` (rows where store == "RVA").

```{r q15, eval=FALSE}
# YOUR CODE + BRIEF NOTE
```

\newpage

# Solutions (not evaluated)

> Instructor note: All solution code chunks are set to `eval=FALSE` so the document will knit without executing solutions.

## 1) Import + types

```{r a1, eval=FALSE}
sales <- sales %>% mutate(date = as.Date(date))
```

---

## 2) Count categories

```{r a2, eval=FALSE}
sales %>% count(category, sort = TRUE)
```

---

## 3) String cleaning

```{r a3, eval=FALSE}
sales %>%
  mutate(
    category_clean = category %>%
      stringr::str_to_lower() %>%
      stringr::str_trim() %>%
      stringr::str_to_title()
  )
```

---

## 4) Percent missing for `revenue`

```{r a4, eval=FALSE}
sales %>% summarise(pct_missing_revenue = mean(is.na(revenue)))
```

---

## 5) Remove rows where `units` OR `revenue` is missing

```{r a5, eval=FALSE}
sales %>% tidyr::drop_na(units, revenue)
```

(Equivalent: `filter(!is.na(units) & !is.na(revenue))`.)

---

## 6) Correct `summarise()` ignoring missingness

```{r a6, eval=FALSE}
sales %>% summarise(avg_rev = mean(revenue, na.rm = TRUE))
```

---

## 7) Average revenue by store

```{r a7, eval=FALSE}
sales %>%
  group_by(store) %>%
  summarise(avg_revenue = mean(revenue, na.rm = TRUE), .groups = "drop")
```

---

## 8) Limitation of only using `na.rm = TRUE`

Example correct responses (any one is sufficient):

- "`na.rm = TRUE` drops missing values only within the calculation; it does not diagnose why values are missing or whether missingness is systematic (which can bias results)."
- "It produces a number but hides the fact that the sample size changed and can mask group differences in missingness."

---

## 9) Join mismatch: fix incompatible key types

A common cause is that one `date` column is `Date` and the other is character. Fix by aligning types before the join.

```{r a9, eval=FALSE}
sales %>%
  mutate(date = as.Date(date)) %>%
  left_join(weather %>% mutate(date = as.Date(date)), by = "date")
```

---

## 10) Dates in `sales` not present in `weather`

```{r a10, eval=FALSE}
sales %>%
  distinct(date) %>%
  anti_join(weather, by = "date")
```

---

## 11) Duplicate key check in `weather`

```{r a11, eval=FALSE}
weather %>% count(date) %>% filter(n > 1)
```

---

## 12) Reshape `wide_temps` to long

```{r a12, eval=FALSE}
temps_long <- wide_temps %>%
  pivot_longer(
    cols = starts_with("Temp_Month_"),
    names_to = "month",
    values_to = "temp",
    names_prefix = "Temp_Month_"
  ) %>%
  mutate(month = as.integer(month))

temps_long
```

---

## 13) Remove pivot artifacts (missing `temp`)

```{r a13, eval=FALSE}
temps_long %>% drop_na(temp)
```

(Equivalent: `filter(!is.na(temp))`.)

---

## 14) Recode with `case_when()`

```{r a14, eval=FALSE}
sales %>%
  mutate(
    store_type = case_when(
      store == "WBG" ~ "Williamsburg",
      store == "RVA" ~ "Richmond",
      TRUE ~ "Other"
    )
  )
```

---

## 15) Grouping explanation + one-row summary

**Explanation:** `group_by(store)` splits the data into groups, so `summarise(n = n())` produces **one row per group** (here, one per store).

A single pipeline returning one row:

```{r a15, eval=FALSE}
sales %>%
  summarise(
    n_total = n(),
    n_wbg   = sum(store == "WBG"),
    n_rva   = sum(store == "RVA")
  )
```
