---
title: "Data Wrangling Pipeline Project"
author: "YOUR NAME"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(janitor)   # optional but helpful: clean_names(), tabyl()
library(skimr)     # optional but helpful: skim()
```

# Overview (read first)

## Purpose
In this project you will build a **reproducible data wrangling pipeline** using the Tidyverse. You will:

1. **Choose a publicly available dataset** of your interest.
2. **Import** the data into R.
3. **Tidy** the dataset into an analysis-ready form (long vs. wide as appropriate).
4. **Diagnose and document missingness**, then make defensible handling decisions.
5. **Aggregate and summarize** the data to answer a focused question.
6. Create **several descriptive plots** and interpret them in writing.

## Time expectations
You have **one week**. The goal is not to do the most complex analysis; it is to demonstrate a clean, defensible wrangling workflow.

## Submission
Submit **one HTML** (knitted from this Rmd) and **your Rmd** source file.
If your dataset is small enough, include it (CSV) in your submission folder; otherwise provide a direct download link and document import steps.

---

# 1. Dataset selection and research question

## 1.1 Data source and citation
Describe your data source (who produced it, where you accessed it, and when you downloaded it). Provide a link.

- Data source:
- Link:
- Date accessed:
- License / terms of use (if stated):

## 1.2 Research question (keep it narrow)
Write **one** clear question your summaries/plots will answer. Examples:
- How does outcome Y vary by group X over time?
- Which categories are most common, and how have they changed?
- What is the distribution of variable X, and how does it differ across groups?


## 1.3 What is one row per what?
State the “grain” of the final tidy dataset you expect (e.g., one row per person per year; one row per county per month).

---

# 2. Import the data

## 2.1 Import code
Use a readr function appropriate to your file format (e.g., `read_csv()`, `read_tsv()`, `read_delim()`).

```{r import}
# Option A: local file in your project folder
# df_raw <- read_csv("YOUR_FILE.csv")

# Option B: direct URL import (works for many raw CSV links)
# df_raw <- read_csv("https://.../YOUR_FILE.csv")

# Option C: multiple files (read and combine)
# files <- list.files("data/", pattern = "\\.csv$", full.names = TRUE)
# df_raw <- files |> map_dfr(read_csv)

#head(df_raw)
```

## 2.2 Quick sanity checks
Document what you see immediately after import.

```{r import_checks}
#glimpse(df_raw)
#summary(df_raw)
```

Write 3–5 bullets:
- What are the key variables?
- Any obvious type issues (numbers imported as text, dates as character)?
- Any obvious missingness?
- Any obviously “messy” structure (wide format, multiple variables in one column, etc.)?

---

# 3. Make the data tidy

## 3.1 Clean names and types
Do cleanup: consistent names, variable values, and correct data types.

```{r clean_names_types}
#library(janitor) # for clean_names() function

#df <- df_raw %>%
#  clean_names() %>%        # snake_case names
#  mutate(
    # Examples (edit to your needs):
    # date = as.Date(date),
    # year = as.integer(year),
    # amount = readr::parse_number(amount),
#  )

#glimpse(df)
```

## 3.2 Subset the data
Create a subset of the data that you will use for this assignment.  
- Select only the columns relevant to the three analyses you will conduct below.
- Select only those observations (i.e., rows) you will be using.

```{r subset_data}
# Example pattern:
# df_subset <- df %>%
#   filter(c(ID,Gender,Dose,Day,Anxiety)) %>%
#   select(Gender=="Male") 

#glimpse(df_tidy)
```

## 3.3 Reshape if needed
If your data is in a “wide spreadsheet” form (e.g., multiple years/months as columns), use `pivot_longer()` to make it tidy. If it’s already tidy, state that clearly and skip.

```{r reshape_if_needed}
# Example pattern:
# df_tidy <- df %>%
#   pivot_longer(
#     cols = starts_with("value_"),
#     names_to = "year",
#     values_to = "value",
#     names_transform = list(year = readr::parse_number)
#   )

df_tidy <- df  # replace with your tidied object if you reshape
glimpse(df_tidy)
```

Explain what you did and why:
- Which columns became key-value pairs (if any)?
- What do rows/columns represent now?

---

# 4. Missingness: diagnose, document, and decide

In a summary: you should quantify where data are missing and then make a decision that you can justify.

## 4.1 Column-level missingness summary
Create a compact table of missing counts and percent missing by column.

```{r missing_by_col}
#missing_by_col <- df_tidy %>%
#  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
#  pivot_longer(
#    cols = everything(),
#    names_to = "column",
#    values_to = "n_missing"
#  ) %>%
#  mutate(
#    pct_missing = n_missing / nrow(df_tidy),
#    n_nonmissing = nrow(df_tidy) - n_missing
#  ) %>%
#  arrange(desc(pct_missing))

#missing_by_col
```

Write 2–4 sentences:
- Which columns have the most missingness?
- Is the missingness “small and ignorable,” or “large and potentially consequential”?

## 4.2 Missingness by subgroup 
If you have a meaningful grouping variable (region, year, category), check whether missingness is concentrated.

```{r missing_by_group}
# Replace group_var with a real variable in your dataset
# missing_by_group <- df_tidy %>%
#   group_by(group_var) %>%
#   summarise(across(everything(), ~ mean(is.na(.x)))) %>%
#   pivot_longer(-group_var, names_to = "column", values_to = "pct_missing") %>%
#   arrange(desc(pct_missing))
#
# missing_by_group
```

Interpretation prompt:
- Do any groups have much higher missingness than others?
- What bias might be introduced if you drop rows with missing values?

## 4.3 Your handling decision (filter vs. fill vs. other)
Choose and apply a missingness strategy for at least **one** important variable.

Options (choose what is defensible for your context):
- **Filter** rows using `drop_na()` (ideally on key variables only).
- **Rule-based fill** using `replace_na()` (only when a replacement is meaningful, e.g., missing count implies 0).
- **Impute** with a simple strategy (mean/median within group) and clearly label it as imputation.
- **Keep missingness** but model/plot it explicitly (e.g., include an “Unknown” category).

```{r missingness_action}
#df_clean <- df_tidy %>%
  # Example 1: Drop rows missing key fields
  # tidyr::drop_na(id, outcome_var)

  # Example 2: Rule-based fill
  # tidyr::replace_na(list(count = 0, category = "Unknown"))

#glimpse(df_clean)
```

Write a short justification:
- Why this approach (filter/fill/impute/keep)?
- What is a plausible bias risk and how did you mitigate it?

---

# 6. Summaries and aggregations

Create at least **two** summary tables:
1. A **descriptive baseline** table (counts, averages, medians).
2. A **question-driven** summary that directly answers your research question.

```{r summaries}
# Example baseline summary
#baseline_summary <- df_final %>%
#  summarise(
#    n_rows = n(),
#    n_cols = ncol(df_final)
    # add context-specific summaries here
#  )
#baseline_summary

# Example grouped summary
# grouped_summary <- df_final %>%
#   group_by(group_var) %>%
#   summarise(
#     n = n(),
#     avg_value = mean(value, na.rm = TRUE)
#   ) %>%
#   arrange(desc(avg_value))
# grouped_summary
```

Write 3–6 sentences interpreting what the summaries show.

---

# 7. Descriptive plots (minimum: 3)

Create at least **three** plots. Your plots should be readable:
- clear title
- labeled axes
- appropriate geometry for variable types

Suggested set:
- Distribution of a numeric variable (histogram/density)
- Comparison across categories (bar chart, boxplot)
- Compare two continuous variables (scatterplot)
- Trend over time (line chart) if you have a time variable

```{r plot1}
# ggplot(df_final, aes(x = numeric_var)) +
#   geom_histogram(bins = 30) +
#   labs(title = "Distribution of ...", x = "...", y = "Count")
```

```{r plot2}
# ggplot(df_final, aes(x = category_var)) +
#   geom_bar() +
#   coord_flip() +
#   labs(title = "Counts by ...", x = "...", y = "Count")
```

```{r plot3}
# ggplot(df_final, aes(x = time_var, y = numeric_var, color = group_var)) +
#   geom_line() +
#   labs(title = "Trend of ...", x = "...", y = "...")
```

For each plot, write 2–3 sentences interpreting the key pattern(s).

---

# 8. Final deliverables checklist

Confirm you have included:

- [ ] Clear data source citation and download/import notes  
- [ ] `glimpse()` and initial diagnostics  
- [ ] A tidy dataset (`df_tidy`) with documented “one row per what”  
- [ ] Missingness summary table (`missing_by_col`) and discussion of missingness risk  
- [ ] A justified missingness decision (filter/fill/impute/keep) and code implementing it  
- [ ] At least two summary tables  
- [ ] At least three descriptive plots with interpretation  
- [ ] Reproducible knit (no manual steps required)

---

# 9. Session info

```{r session_info}
sessionInfo()
```
