---
title: "Test: Introduction to Programming in R (Base R Only)"
author: ""
date: ""
output:
  pdf_document:
    toc: false
    number_sections: false
---

> **Instructions:** Answer all 30 questions.  
> Each question is one of the following types:  
> (1) **Describe output/behavior** of R code (focus on type/structure/what prints),  
> (2) **Find and fix an error**,  
> (3) **Write R code** to achieve a specific goal,  
> (4) **Short debugging/reasoning** (brief and specific).  
>
> Do **not** do hand calculations; when appropriate, describe results in terms of **type, length, names, or which branch runs**.

# Reference objects (used across questions)

You may assume the following objects exist. (They are provided for context; you do **not** need to run them during the test.)

```{r reference_objects, eval=FALSE}
# Basic vectors
nums <- c(-2, 0, 3, NA, 10)
letters_vec <- c("alpha", "beta", "gamma")

# Simple data frame (base R)
df <- data.frame(
  id = 1:6,
  group = c("A", "A", "B", "B", "B", "A"),
  score = c(10, NA, 8, 9, NA, 7),
  stringsAsFactors = FALSE
)

# A simple function with one input (used for behavior questions)
make_label <- function(x) {
  if (is.na(x)) {
    "missing"
  } else if (x >= 9) {
    "high"
  } else {
    "low"
  }
}

# A simple function with multiple outputs via a list
summarize_missing <- function(x) {
  list(
    n_total = length(x),
    n_missing = sum(is.na(x)),
    any_missing = any(is.na(x))
  )
}
```

\newpage

# Part A — Describe the output/behavior (Questions 1–12)

## 1) Output/behavior: assignment vs printing

```{r q1, eval=FALSE}
x <- 5
x
y <- x + 1
```

Which lines print output to the console, and what is printed (in general terms)?

---

## 2) Output/behavior: `class()` vs `typeof()`

```{r q2, eval=FALSE}
class(nums)
typeof(nums)
```

What is the difference between what these two functions report (conceptually)?

---

## 3) Output/behavior: missingness propagation

```{r q3, eval=FALSE}
mean(nums)
```

Does this return `NA` or a numeric value? Why?

---

## 4) Output/behavior: removing missing values

```{r q4, eval=FALSE}
mean(nums, na.rm = TRUE)
```

How does this behavior differ from Question 3?

---

## 5) Output/behavior: `ifelse()` is vectorized

```{r q5, eval=FALSE}
ifelse(is.na(nums), "MISSING", "PRESENT")
```

Describe the output’s **type** and **length**.

---

## 6) Output/behavior: vectorized function call (conceptual)

```{r q6, eval=FALSE}
sapply(df$score, make_label)
```

What is returned (type/length), and why does this work even though `make_label()` uses `if`?

---

## 7) Output/behavior: list names

```{r q7, eval=FALSE}
out <- summarize_missing(nums)
names(out)
```

What does `names(out)` return?

---

## 8) Output/behavior: list element extraction

```{r q8, eval=FALSE}
out$any_missing
```

What is the type of this value (logical/numeric/character/list)?

---

## 9) Output/behavior: base R subsetting by condition

```{r q9, eval=FALSE}
df[df$group == "B", ]
```

What is returned (data type/structure), and how many rows should it contain?

---

## 10) Output/behavior: selecting a single column

```{r q10, eval=FALSE}
df[df$group == "B", "score"]
```

What is returned (a vector or a data frame), and what is its length?

---

## 11) Output/behavior: `if` / `else` branch (no arithmetic)

```{r q11, eval=FALSE}
z <- NA
if (is.na(z)) {
  print("missing")
} else {
  print("present")
}
```

What is printed?

---

## 12) Output/behavior: `for` loop printing

```{r q12, eval=FALSE}
for (g in unique(df$group)) {
  print(paste("Group:", g))
}
```

How many lines are printed, and what is their general format?

\newpage

# Part B — Find and fix the error (Questions 13–22)

## 13) Error: `if` condition must be length 1

```{r q13, eval=FALSE}
x <- c(TRUE, FALSE)
if (x) print("yes")
```

Why is this an error, and provide one correct fix.

---

## 14) Error: using `=` instead of `==` in a condition

```{r q14, eval=FALSE}
x <- 3
if (x = 3) print("equal")
```

Identify the mistake and provide the corrected line.

---

## 15) Error: missing comma in data frame indexing

```{r q15, eval=FALSE}
df[df$group == "A"]
```

Explain what is wrong and provide corrected code that returns **all columns** for group A.

---

## 16) Error: wrong argument name

```{r q16, eval=FALSE}
mean(nums, na_rm = TRUE)
```

What is the error, and what is the corrected function call?

---

## 17) Error: unmatched parentheses

```{r q17, eval=FALSE}
print(paste("ID:", df$id[1])
```

Fix the syntax.

---

## 18) Error: function braces/structure

```{r q18, eval=FALSE}
first_last <- function(x) 
  first <- x[1]
  last <- x[length(x)]
  list(first = first, last = last)
}
```

Identify the syntax problem and provide a corrected function definition.

---

## 19) Error: infinite loop risk

```{r q19, eval=FALSE}
i <- 1
while (i <= 3) {
  print("looping")
}
```

What is the problem, and what single line must be added (or modified) so the loop terminates?

---

## 20) Error/logic: returning multiple values

```{r q20, eval=FALSE}
two_outputs <- function(x) {
  a <- head(x, 1)
  b <- tail(x, 1)
  a
  b
}
```

What does this function actually return, and how should it be rewritten to return **both** values?

---

## 21) Error: calling a function with a misspelled argument

```{r q21, eval=FALSE}
proportion_above <- function(x, cutoff = 0) {
  mean(x > cutoff, na.rm = TRUE)
}

proportion_above(nums, cutof = 3)
```

Why does the call fail, and what is the corrected call?

---

## 22) Error: mixing character and numeric in arithmetic

```{r q22, eval=FALSE}
a <- "10"
a + 1
```

Describe the error in plain language and give one line of code to fix it.

\newpage

# Part C — Write R code (Questions 23–30)

## 23) Write code: define a simple function

Write a function `is_missing(x)` that returns `TRUE` if `x` is `NA` and `FALSE` otherwise.

```{r q23, eval=FALSE}
# YOUR CODE
```

---

## 24) Write code: safe mean (behavior-focused)

Write a function `safe_mean(x)` that returns:
- `NA_real_` if **all** values of `x` are missing, otherwise
- `mean(x, na.rm = TRUE)`

```{r q24, eval=FALSE}
# YOUR CODE
```

---

## 25) Write code: default arguments

Write a function `clip_to_range(x, low = 0, high = 7)` that clips values below `low` up to `low`, and values above `high` down to `high`.

```{r q25, eval=FALSE}
# YOUR CODE
```

---

## 26) Write code: named argument call

Show a call that clips `c(-2, 1, 3, 10)` to the range **0 to 7** using **named arguments**.

```{r q26, eval=FALSE}
# YOUR CODE
```

---

## 27) Write code: return a named list

Write a function `descriptive_flags(x)` that returns a named list with:
- `any_missing` (logical)
- `all_missing` (logical)
- `n_missing` (numeric)

```{r q27, eval=FALSE}
# YOUR CODE
```

---

## 28) Write code: use `ifelse()` to label a vector

Write one line of code that creates a character vector `status` from `nums` where missing values become `"missing"` and non-missing become `"present"`.

```{r q28, eval=FALSE}
# YOUR CODE
```

---

## 29) Write code: `for` loop that builds a result vector

Write code that loops over `nums` and produces a new vector `labels` such that:
- `"missing"` if the element is `NA`
- `"nonmissing"` otherwise

(You may use preallocation with `character(length(nums))`.)

```{r q29, eval=FALSE}
# YOUR CODE
```

---

## 30) Creative debugging prompt: “read your own code”

You are given the following buggy function. Write **two** lines you would add (e.g., `print()` statements) to help you debug what values are flowing through the function, and briefly state what you are trying to learn from those prints.

```{r q30, eval=FALSE}
mystery <- function(x) {
  y <- x[x > 0]
  if (length(y) == 0) return(NA_real_)
  mean(y, na.rm = TRUE)
}
# Add two debugging lines (do not rewrite the whole function)
```
