---
title: "Programming II Assignment: Loops & Flow Control"
subtitle: "Data Science for the Psychological Sciences"
author: "Student Name"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: false
    toc_depth: 2
    number_sections: false
  pdf_document:
    toc: false
    number_sections: false
    latex_engine: xelatex
geometry: margin=1in
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
options(htmltools.dir.version = FALSE)
```

# Overview and instructions

## Purpose
This assignment is focused on:

- 

## What to submit
Submit **two files**:

1. `Lastname_Firstname_Programming_II_Assignment.Rmd`
2. The knitted output: `Lastname_Firstname_Programming_II_Assignment.html`

Your report must **knit without errors**.

**General expectations**
- Your functions should be readable and use clear names.
- You must demonstrate tests (function calls) that show your functions work.

---

# Part 1: Variables and Basic Operations

## Problem 1: Area of a Rectangle

Create two variables, `width` and `height`, and assign them numeric
values (e.g., 10 and 5). Calculate the area of the rectangle
(`width * height`) and store it in a variable called `area`. Print the
`area`.

```{r  p1_solution}
# Your code here
```

## Problem 2: Full Name Greeting

Create a variable `first_name` and a variable `last_name`. Use the
`paste()` function to combine them into a single string that says
"Hello, [First Name] [Last Name]!" and print the result.

```{r p2_solution}
# Your code here
```

## Problem 3: Is the Number Even?

Write a **function** that takes a number as input and determines whether
or not that number is odd or even, returning the logical `TRUE` if the
number is even or `FALSE` if the number is odd. (Hint: the modulus
operator `%%` might come in handy here). Call the function twice after
you've created it to demonstrate that it works.

```{r p3_solution}
# Your code here
```

------------------------------------------------------------------------

# Part 2: Vectors and Logical Operators

## Problem 4: Monthly Expenses

Create a data frame named `expenses` that contains the following
"expense categories": rent, groceries, utilities, transportation, and
entertainment with the corresponding "amounts": `800` `250` `120` `300` `150`. Use logical indexing and the `sum()`
function to compute monthly expenses *excluding* entertainment.

```{r p4_solution}
# Your code here
```

## Problem 5: Accessing Vector Elements

Using the `expenses` vector from the previous problem, write code to
access and print only the expense for groceries (the 2nd element) and
the expense for entertainment (the 5th element).

```{r p5_solution}
# Your code here
```

## Problem 6: Which Expenses are High?

Using the `expenses` vector, write a line of code that returns a logical
vector (`TRUE`/`FALSE` values) indicating which expenses are greater
than `200`.

```{r p6_solution}
# Your code here
```

## Problem 7: Checking Expenses in a Data Frame

Create a data frame with the following column names: (1) Month (2) Rent,
(3) Groceries, (4) Travel, and (5) Holiday. The values in the month
column should include January-December Use any reasonable \$ values for
Rent, Groceries, and Travel each of the six months. The Holiday column
should contain logical values (i.e., TRUE or FALSE) indicating whether
or not the month contains a major Holiday typically associated with travel (this may vary from person to
person).

Assuming that travel costs are likely to be higher in months where there
is a holiday, use indexing and the `mean()` function to compute the
average travel expense in months where there is a holiday and in months
that do not include a holiday.

```{r p6b_solution}
# Your code here
```

------------------------------------------------------------------------

# Part 3: Flow Control and Loops

## Problem 8: Movie Rating Guide
Imagine a movie rating system. Write an `if`/`else if`/`else` statement that checks a variable `rating` and
prints: - "Excellent!" if the rating is 8 or higher. - "Good" if the
rating is between 5 and 8 (exclusive of 8). - "Poor" for anything else.
Test your code with `rating <- 8`, `rating<-6`, and `rating<-2`.

```{r p7_solution}
rating <- 9
# Your if/else if/else statement here
```

## Problem 9: Temperature Conversion Loop
You have a vector of temperatures in Celsius. Use a `for` loop to iterate through each
temperature, convert it to Fahrenheit, and print the result. The formula
is `(Celsius * 9/5) + 32`.

```{r p8_solution}
celsius_temps <- c(0, 15, 30, -5)

# Your for loop here
```

## Problem 10: Reaching a Savings Goal 

You want to save \$50. You start with \$10 and save \$5 each week. Use a `while` loop to determine how
many weeks it will take to reach your goal. Inside the loop, you should
increment the weeks and add to your savings. Print the final number of
weeks.

```{r p9_solution}
savings <- 10
goal <- 50
weeks <- 0

# Your while loop here

print(paste("It will take", weeks, "weeks to reach the goal."))
```


## Problem 11: Create a Subset Data Frame 
The `mtcars` dataset is already built into R. For this first step, create a *new* data frame
called `my_cars` that contains only four columns from `mtcars`: `mpg`
(Miles Per Gallon), `hp` (Horsepower), `wt` (Weight), and `cyl`
(Cylinders). (Hint: `new_df <- old_df[, c("col1", "col2")]`). Print the
first few rows of your new data frame using `head()`.

```{r p10_solution}
# Your code here
```

## Problem 12: Find the Highest Horsepower

Using the `my_cars` data frame you just created, use a `for` loop to find the highest horsepower of any car in your subset.

```{r p11_solution}
# Your code here
```

## Problem 13: Calculate Average Horsepower
Write a *function* that will calculate the average horsepower from the `hp` column of your `my_cars` data frame, separately for the 4-cylinder and 8-cylinder cars (indicated in the `cyl` column) and return the two means in a list object.

Once you have it working, test it on the original mt_cars data set.

```{r p12_solution}
# Your code here
```

------------------------------------------------------------------------

# Part 4: Bringing It All Together

## Problem 14: Interactive Number Guessing Game
This problem will test your ability to combine `readline()`, a `while` loop, and
`if/else if/else` statements to create a simple interactive program.

**Your Goal:** Write a script that plays a number guessing game with the
user.

**Instructions:**

1.  **Set a secret number.** Create a variable named `secret_number` and
    assign it an integer value between 1 and 100 (e.g., `42`).

2.  **Create a loop.** 
  - HINT: Use `while (TRUE)` to create a loop that will run forever until you explicitly tell it to stop.
  - Set a maximum number of "guesses" and loop until that maximum number of guesses has been met.

3.  **Get user input.** Inside the loop, use `readline()` to prompt the
    user to guess a number. Store their input in a variable.

4.  **Convert the input.** Remember that `readline()` returns a
    character string. You must convert the user's input into a number
    using `as.numeric()` before you can compare it to your
    `secret_number`.

5.  **Check the guess.** Use an `if/else if/else` statement with logical
    operators to check the user's guess against the `secret_number`:

    -   If the guess is **less than** the secret number, print "Too low!
        Guess again."
    -   If the guess is **greater than** the secret number, print "Too
        high! Guess again."
    -   If the guess is **equal to** the secret number, print a success
        message like "Correct! You guessed the number!" and then use the
        `break` command to exit the `while` loop.
    -   Keep track of how many guesses the user takes using a
        `guess_counter` variable that is initially set to 0. Inside the
        loop, add 1 to the counter for each guess. When the user guesses
        correctly, include the number of guesses in the success message
        (e.g., "Correct! You guessed the number in 5 tries!").

```{r p16_solution}
# 1. Set the secret number
secret_number <- 42

print("I'm thinking of a number between 1 and 100.")

# 2. Start the loop{
  
#}
```

**Problem 15: The Fuel Efficiency Expert**

**Objective:** Your goal is to analyze the `mtcars` dataset to
automatically generate a performance summary. You will write a script
that iterates through the different cylinder groups (4, 6, and 8
cylinders), calculates the average fuel efficiency for each group, and
then prints a customized evaluation based on that average.

This problem will require you to use a `for` loop, perform logical
indexing to filter a data frame, and use `if`/`else if`/`else`
statements to make decisions.

**Instructions:**

1.  **Create a vector of cylinder groups.** Make a vector named
    `cylinder_groups` that contains the unique cylinder values you want
    to analyze: `c(4, 6, 8)`.

2.  **Start a `for` loop.** Write a `for` loop that iterates through
    each value in your `cylinder_groups` vector.

3.  **Inside the loop, filter the data.** For each cylinder count in the
    loop, use **logical indexing** to create a temporary subset of the
    `mtcars` data frame that contains *only* the cars with that number
    of cylinders.

4.  **Calculate the average MPG.** For the temporary subset you just
    created, calculate the mean of the `mpg` column. Store this value in
    a variable (e.g., `avg_mpg`).

5.  **Use conditional logic to evaluate efficiency.** Write an
    `if`/`else if`/`else` statement that checks the `avg_mpg` you just
    calculated:

    -   If `avg_mpg` is greater than 25, print a message like:
        `"Cars with [X] cylinders are highly fuel-efficient."`

    -   If `avg_mpg` is between 18 and 25 (inclusive), print a message
        like: `"Cars with [X] cylinders have moderate fuel efficiency."`

    -   Otherwise (if `avg_mpg` is less than 18), print a message like:
        `"Cars with [X] cylinders are not fuel-efficient."`

6.  Make sure to replace `[X]` in your print statements with the actual
    cylinder count from the current loop iteration. The `paste()`
    function will be very helpful here.

**Starter Code:**

```{r efficiency_expert_solution}
# 1. Create a vector of cylinder groups to loop through
cylinder_groups <- c(4, 6, 8)

# 2. Start the for loop
for (cyl in cylinder_groups) {
  
  # 3. Use logical indexing to create a subset for the current cylinder group
  # (Your code here)
  
  # 4. Calculate the average MPG for the subset
  # (Your code here)

  # 5. Use if/else if/else to check the average MPG and print a summary
  # (Your code here)

}
```

**Expected Output:** When you run your completed code, you should see
something similar to the following output in your console:

```         
[1] "Cars with 4 cylinders are highly fuel-efficient."
[1] "Cars with 6 cylinders have moderate fuel efficiency."
[1] "Cars with 8 cylinders are not fuel-efficient."
```

------------------------------------------------------------------------

**Problem 16: The Plant Species Analyst**

**Objective:** You are an assistant to a botanist who is studying iris
flowers. The botanist has a theory that the species with the largest
average petal area is the most resilient. Your task is to analyze the
built-in `iris` dataset to identify which of the three species has the
largest average petal area.

This problem will require you to get unique values for a column using
the `unique()` function, loop through those values using a `for` loop,
perform logical indexing, create a new calculated value, and use an `if`
statement with "tracker" variables to find the maximum value across all
categories.

**Instructions:**

1.  **Get the unique species.** First, create a vector called
    `species_names` that contains the unique species from the
    `iris$Species` column. The `unique()` function will be perfect for
    this.

2.  **Initialize "tracker" variables.** Before you start the loop, you
    need variables to keep track of the winner. Create two variables:

    -   `max_avg_petal_area` and set it to `0`.
    -   `winning_species` and set it to an empty string `""`.

3.  **Start a `for` loop.** Write a `for` loop that iterates through
    each name in your `species_names` vector.

4.  **Inside the loop, filter the data.** For each species, use
    **logical indexing** to create a temporary subset of the `iris` data
    frame.

5.  **Calculate petal area.** For your temporary subset, calculate the
    average petal area. You can estimate this as
    `mean(subset$Petal.Length * subset$Petal.Width)`. Store this value
    in a variable like `current_avg_area`.

6.  **Use an `if` statement to check for a new winner.** Compare the
    `current_avg_area` to your `max_avg_petal_area` tracker.

    -   If the current species' average petal area is greater than the
        max area found so far, you have a new winner! You must update
        both of your tracker variables:
        -   Set `max_avg_petal_area` to the `current_avg_area`.
        -   Set `winning_species` to the name of the current species.

7.  **After the loop, print the final result.** Once the loop has
    finished checking all the species, your tracker variables will hold
    the final answer. Print a single, formatted summary sentence
    announcing the winner. The `paste()` function will be helpful.

**Starter Code:**

```{r species_analyst_solution}
# 1. Get the unique species from the iris dataset
species_names <- unique(iris$Species)

# 2. Initialize tracker variables
max_avg_petal_area <- 0
winning_species <- ""

# 3. Start the for loop to iterate through each species
for (species in species_names) {
  
  # 4. Use logical indexing to create a subset for the current species
  # (Your code here)

  # 5. Calculate the average petal area for the subset
  # (Your code here)

  # 6. Use an if statement to check if this species is the new winner
  # (Your code here)
  
}

# 7. Print the final result after the loop is finished
# (Your code here)
```

**Expected Output:** When your code is run correctly, it should produce
a single summary line that looks like this:

```         
[1] "The winning species is virginica with an average petal area of 10.88"
```

*(Note: Your exact number might vary slightly depending on rounding,
which is fine.)*

# Acknowledgments (required only if you used outside help)

List any people, websites, or tools you used. Be specific enough that your work is transparent.

-
