% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
  11pt,
]{article}
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage[]{libertine}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage[]{natbib}
\bibliographystyle{apsr}
\usepackage{array}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[table]{xcolor}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{calc}
\usepackage{tabularx}
\usepackage{fontawesome}
\usepackage[para,online,flushleft]{threeparttable}
\usepackage{wrapfig}
\captionsetup[wrapfigure]{justification=raggedright,singlelinecheck=false}
\usepackage{hyperref}
\hypersetup{colorlinks=true, urlcolor=blue}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Finding Your Dataset},
  pdfauthor={Data Science for the Psychological Sciences},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Finding Your Dataset}
\author{Data Science for the Psychological Sciences}
\date{}

\begin{document}
\maketitle

\section{Purpose}\label{purpose}

In data science, the single most important early skill is learning to
locate, evaluate, and justify a dataset that is appropriate for a
research question. This assignment asks you to identify a public
dataset, assess its quality and constraints, and propose a feasible
analysis plan.

\textbf{Student tasks and deliverables}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Find three public datasets and briefly record:
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Dataset name + source
\item
  Topic/domain (psychology, mental health, public health, politics,
  etc.)
\item
  Unit of analysis (people, countries, counties, survey responses, etc.)
\item
  Approximate sample size and key variables
\item
  Access method (direct download vs registration; file format)

  \begin{itemize}
  \tightlist
  \item
    Provide link if available
  \end{itemize}
\item
  Documentation availability (codebook/data dictionary)
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Select one of the datasets and write a short proposal with the
  following sections:
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Dataset overview
\item
  Source, citation, licensing/terms-of-use
\item
  Unit of analysis, time span, sampling frame
\item
  Key variables (at least 6) and why they matter to you
\item
  Documentation and data quality

  \begin{itemize}
  \tightlist
  \item
    Do the data look interpretable? Is there a codebook?
  \end{itemize}
\item
  3 potential research questions
\end{itemize}

\section{Curated public dataset list}\label{curated-public-dataset-list}

Choose three of the datasets below \textbf{OR} find your own.

\subsection{Mental health / psychiatry / substance
use}\label{mental-health-psychiatry-substance-use}

NSDUH (National Survey on Drug Use and Health) --- substance use, mental
health indicators, treatment utilization (public-use files).
\href{https://www.samhsa.gov/data/data-we-collect/nsduh-national-survey-drug-use-and-health/datafiles}{SAMHSA}

BRFSS (Behavioral Risk Factor Surveillance System) --- U.S. health
behaviors, preventive care, chronic conditions; huge sample; state-level
focus. \href{https://www.cdc.gov/brfss/annual_data/annual_data.htm}{CDC}

NHIS (National Health Interview Survey) --- health status, access,
behaviors; rich health and demographic content.
\href{https://www.cdc.gov/nchs/nhis/documentation/index.html}{CDC}

NHANES --- interviews + exams + lab measures; includes mental health and
health behavior modules across waves.
\href{https://wwwn.cdc.gov/nchs/nhanes/Default.aspx}{CDC}

YRBSS (Youth Risk Behavior Surveillance System) --- adolescent risk
behaviors (sleep, substance use, mental health--related items vary by
year). \href{https://www.cdc.gov/yrbs/data/index.html}{CDC}

\subsection{Psychology and behavioral science (open research
data)}\label{psychology-and-behavioral-science-open-research-data}

Collaborative Psychiatric Epidemiology Surveys (CPES) - Harmonized U.S.
mental health surveys (includes NCS-R plus major race/ethnicity-focused
components) used for studying prevalence, correlates, and treatment
patterns.
\href{https://www.icpsr.umich.edu/web/ICPSR/studies/20240/versions/V3}{icpsr.umich.edu}

NESARC (National Epidemiologic Survey on Alcohol and Related Conditions;
waves and NESARC-III) - Very strong for comorbidity (substance use +
mood/anxiety/personality disorders, etc.) and demographic correlates;
documentation and access pathways are well established.
\href{https://catalog.data.gov/dataset/national-epidemiologic-survey-on-alcohol-and-related-conditions-nesarcwave-1-20012002-and-}{Data.gov}

NSDUH (National Survey on Drug Use and Health) public-use files - Annual
U.S. survey with mental illness indicators, suicidal ideation/behavior
items (varies by year), and treatment variables---excellent for
population trends and disparities.
\href{https://www.samhsa.gov/data/data-we-collect/nsduh-national-survey-drug-use-and-health/datafiles}{SAMHSA}

\subsection{Politics and public
opinion}\label{politics-and-public-opinion}

ANES (American National Election Studies) --- U.S. political attitudes,
vote choice, ideology, demographics.
\href{https://electionstudies.org/data-center/}{ANES}

CES/CCES (Cooperative Election Study) --- large U.S. survey; common
content + modules; excellent for modeling turnout/attitudes.
\href{https://cces.gov.harvard.edu/}{CES/CCES}

GSS (General Social Survey) --- long-running U.S. social attitudes,
demographics, and repeated items.
\href{https://gss.norc.org/get-the-data.html}{GSS}

Pew Research Center datasets --- many public microdata releases after a
period of time; great topical coverage.
\href{https://www.pewresearch.org/download-datasets}{Pew Research
Center}

\subsection{Cross-national attitudes, values, and
governance}\label{cross-national-attitudes-values-and-governance}

European Social Survey (ESS) --- cross-national attitudes/behavior;
strong documentation; repeated rounds.
\href{https://www.europeansocialsurvey.org/data-portal}{European Social
Survey}

World Values Survey (WVS) --- global values/attitudes across waves;
broad psychosocial variables.
\href{https://www.worldvaluessurvey.org/WVSContents.jsp}{worldvaluessurvey.org}

V-Dem (Varieties of Democracy) --- country-year democracy
indicators/indices; good for political context variables.
\href{https://v-dem.net/data/the-v-dem-dataset/}{V-Dem}

\subsection{Population, inequality, and social determinants (useful for
psych/public health
context)}\label{population-inequality-and-social-determinants-useful-for-psychpublic-health-context}

IPUMS USA (ACS/Census harmonized extracts) --- demographics, income,
education, migration; great for social determinants questions (requires
free registration and extract request).
\href{https://usa.ipums.org/usa/data.shtml}{IPUMS USA}

DHS Program (Demographic and Health Surveys) --- global health and
demographics; requires account approval; strong for public health and
international comparisons. \href{https://www.dhsprogram.com/data/}{DHS
Program}

\end{document}
