---
title: "Cluster Analysis Research Report"
author: "Paul Kieffaber"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: false
    df_print: paged
    code_download: true
subtitle: An Investigation into the StarWars Data
---
```{css, echo=FALSE}
h1, h4 {
  text-align: center;
}
```

```{r setup, include=FALSE}
# This global option ensures NO code appears in the final file
knitr::opts_chunk$set(
  echo = FALSE,       # Hide code
  message = FALSE,    # Hide package loading messages
  warning = FALSE,    # Hide warnings
  fig.align = "center" # Center all graphs
)
```

**Goal.** Pick a publicly available dataset, ask a concrete question about how observations might group together, then (1) determine a sensible number of clusters using the elbow and silhouette methods, (2) compare K‑means, Hierarchical (Ward.D2), and DBSCAN, (3) select and justify a final solution, and (4) compare clusters on at least one validation variable not used to form clusters.

# The Star Wars Dataset

The starwars dataset (included in the dplyr package) contains information on 87 characters from the Star Wars universe.

- Some Key variables: 
```{r}
library(dplyr)
glimpse(starwars)
```
  
**Research Question**
Are there unique groupings of Star Wars characters based on their physical characteristics?

# Pre‑process & feature selection

- There are three continuous variables that can be used for clustering:
  - Height
  - Mass
  - Birth_year

```{r}
library(ggplot2)
library(patchwork)
# Plot
p1<-ggplot(data=starwars, aes(x=1,y=height)) +
    geom_violin(alpha=.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme(
      legend.position="none",
      plot.title = element_text(size=12),
      axis.text = element_text(size=12)
    ) +
    ggtitle("Character Height") +
    xlab("")

p2<-ggplot(data=starwars, aes(x=1,y=mass)) +
    geom_violin(alpha=.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme(
      legend.position="none",
      plot.title = element_text(size=12),
      axis.text = element_text(size=12)
    ) +
    ggtitle("Character Mass") +
    xlab("")

p3<-ggplot(data=starwars, aes(x=1,y=birth_year)) +
    geom_violin(alpha=.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme(
      legend.position="none",
      plot.title = element_text(size=12),
      axis.text = element_text(size=12)
    ) +
    ggtitle("Character Birth Year") +
    xlab("")

p1+p2+p3
```

Looking at these variables, it appears that Mass>500 and Birth_year>250 will remove outliers in those variables (see filtered data below).

```{r}
starwars_clean<-starwars%>%
  filter(mass<500)%>%
  filter(birth_year<250)

p1<-ggplot(data=starwars_clean, aes(x=1,y=height)) +
    geom_violin(alpha=.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme(
      legend.position="none",
      plot.title = element_text(size=12),
      axis.text = element_text(size=12)
    ) +
    ggtitle("Character Height") +
    xlab("")

p2<-ggplot(data=starwars_clean, aes(x=1,y=mass)) +
    geom_violin(alpha=.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme(
      legend.position="none",
      plot.title = element_text(size=12),
      axis.text = element_text(size=12)
    ) +
    ggtitle("Character Mass") +
    xlab("")

p3<-ggplot(data=starwars_clean, aes(x=1,y=birth_year)) +
    geom_violin(alpha=.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme(
      legend.position="none",
      plot.title = element_text(size=12),
      axis.text = element_text(size=12)
    ) +
    ggtitle("Character Birth Year") +
    xlab("")

p1+p2+p3
```

Data for clustering were then normalized (see below).

```{r}
clusterdata<-starwars_clean%>%
  select(c(height,mass,birth_year))%>%
  scale()

head(clusterdata)
```

# Decide the number of clusters with elbow & silhouette
```{r,message = FALSE, warning = FALSE}
library(factoextra)
# Find optimal K using the Elbow method
fviz_nbclust(clusterdata, kmeans, method = "wss") +
  labs(subtitle = "Elbow Method")
```

```{r,message = FALSE, warning = FALSE}
# Find optimal K using the Silhouette method
fviz_nbclust(clusterdata, kmeans, method = "silhouette") +
  labs(subtitle = "Silhouette Method")
```
The elbow plot and silhouette method appear to be in agreement that four clusters may be the most appropriate (i.e., $k=4$).

# Fit and visualize three methods

## K-means Clustering (Hartigan-Wong method)

The PCA plot shows good separation of the clusters with the two PCA dimensions accounting for more than 92% of the variabiltiy in the data.

```{r,message = FALSE, warning = FALSE}
# Set seed for reproducibility
set.seed(123)

# Run K-Means
# nstart=25 will run the algorithm 25 times with different
# random starting points and pick the best one.
km.res <- kmeans(clusterdata, centers = 4, nstart = 25, algorithm = c("Hartigan-Wong"))

# Visualize the clusters
# fviz_cluster will perform PCA (Principle Component Analysis)
# to show the 4-dimensional data in 2 dimensions.
kmplot<-fviz_cluster(km.res, data = clusterdata,
             ellipse.type = "convex",
             geom = "point",
             ggtheme = theme_bw())

kmplot
```
CLUSTER SUMMARY:
```{r, message = FALSE, warning = FALSE}
# Add cluster assignments to the original data
clusterdata <- as.data.frame(clusterdata) %>%
  mutate(KMcluster = km.res$cluster)

# Calculate the mean of each variable for each cluster
cluster_profiles_km <- clusterdata %>%
  group_by(KMcluster) %>%
  summarise(
    n = n(),
    Height_mean = mean(height),
    Mass_mean = mean(mass),
    Age_mean = mean(birth_year)
  )

print(cluster_profiles_km)
```

It looks like most of my clusters are pretty small.
***Interpretation:***

* Cluster 1:  Short, light, and young.
* Cluster 2:  Tall, heavy, average age
* Cluster 3:  Tall, average mass, very old
* Cluster 4:  Average height, average mass, average age

## Heirarchical Clustering

```{r, message = FALSE, warning = FALSE}
# 1. Calculate the distance matrix
# This calculates the distance between all 50 states
d <- dist(clusterdata, method = "euclidean")

# 2. Run hierarchical clustering
# We feed the distance matrix 'd' into hclust
# We specify the "ward.D2" linkage method, which minimizes
# the increase in variance at each merge.
hc.res <- hclust(d, method = "ward.D2")
# 3. Plot the dendrogram
fviz_dend(hc.res, cex = 0.7,  # cex = label size
          k = 4, # "Cut" the tree into 4 clusters
          rect = TRUE, # Add rectangles around the 4 clusters
          k_colors = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07"),
          main = "Hierarchical Clustering Dendrogram")
```
```{r, message = FALSE, warning = FALSE}
# Cut the tree into 4 groups
clusters_hc <- cutree(hc.res, k = 4)

# Visualize the clusters on a scatter plot
hplot<-fviz_cluster(list(data = clusterdata, cluster = clusters_hc),
             ellipse.type = "convex",
             geom = "point",
             ggtheme = theme_bw())

hplot
```
CLUSTER SUMMARY:
```{r, message = FALSE, warning = FALSE}
# Add cluster assignments to the original data
clusterdata <- as.data.frame(clusterdata) %>%
  mutate(KMcluster = clusters_hc)

# Calculate the mean of each variable for each cluster
cluster_profiles_km <- clusterdata %>%
  group_by(KMcluster) %>%
  summarise(
    n = n(),
    Height_mean = mean(height),
    Mass_mean = mean(mass),
    Age_mean = mean(birth_year)
  )

print(cluster_profiles_km)
```
The clusters identified using hierarchical clustering are very similar to those using K-means clustering.

## Density Based Clustering (DBSCAN)
```{r, message = FALSE, warning = FALSE}
library(dbscan)
# minPts = k+1. If we want k=4 neighbors, minPts = 5.
# This plots the distance to the 4th nearest neighbor for all 50 states
kNNdistplot(clusterdata, k = 4)
abline(h = 1.25, lty = 2, col = "red") # lty=2 is a dashed line
```
- The nearest neighbor distance plot indicates an epsilon of about 1.25 with k=4, so I'll use epsilon=1.24 with minpnts=5 (k+1)

```{r, message = FALSE, warning = FALSE}
# Run DBSCAN
db.res <- dbscan(clusterdata, eps = 1.25, minPts = 5)

# Print the results
print(db.res)
```
Ok, so the DBSCAN method suggests there are no natural clusters, that everyone should be in the same cluster with 9 outliers.

```{r, message = FALSE, warning = FALSE}
# Visualize, note how '0' (black points) are the outliers
dbplot<-fviz_cluster(db.res, data = clusterdata,
             ellipse = TRUE,
             geom = "point",
             ggtheme = theme_bw())

dbplot
```
This solution seems strange.  When reduced to 2 dimensions, two of the outliers fall within the cluster.

### Comparison
```{r}
library(patchwork)
kmplot+hplot+dbplot
```
# Pick a final solution & justify
Kmeans & Hierarchical were nearly identical.  So I'm going to pick the k-means for validation

# Validate clusters on an external variable

```{r}
# 1. Add your cluster data to the main data frame
# (I'll assume your clusterdata$KMcluster vector is correct)
starwars_with_clusters <- starwars_clean %>%
  mutate(KMcluster = factor(clusterdata$KMcluster)) # Using factor() is good practice

# 2. Count the occurrences of each skin_color within each cluster
plot_data <- starwars_with_clusters %>%
  count(KMcluster, species) # This is the magic!

# 3. Plot the result as a stacked bar chart
library(ggplot2)
ggplot(plot_data, aes(x = KMcluster, y = n, fill = species)) +
  geom_col() + # geom_col is for bar charts where 'y' is a value
  labs(
    title = "Distribution of Species by Cluster",
    x = "K-Means Cluster",
    y = "Count of Characters"
  )
```
In order to validate the clustering, I decided to test whether certain species were more or less likely to be in a cluster using a chi-squared test of association.

```{r}
# 1. Run the test and save the results
final_data <- starwars_clean %>%
  mutate(KMcluster = factor(clusterdata$KMcluster))

# 2. Create the table
contingency_table <- table(final_data$KMcluster, final_data$species)

# 3. Run the Chi-Squared test
chi_test_results <- chisq.test(contingency_table)
print(chi_test_results)
```
The results indicate that the star wars characters are NOT clustered by physical characteristics into distinct species.  The Chi-squared test did not reach statistical significance, indicating that clustering on physical attributes did not relate to species.
