---
title: "Cluster Analysis — Lecture Summary"
subtitle: "From Data to Discovery: Data Science for the Psychological Sciences"
author: "Your Name"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  slidy_presentation:
    incremental: true
    self_contained: true
bibliography: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE, eval = FALSE, fig.width = 7, fig.height = 4.5)
```

# Learning Goals

- Distinguish **cluster analysis** from **PCA/Factor Analysis**.
- Understand why **scaling** matters and how **distance metrics** work.
- Compare three families of algorithms: **K-means**, **Hierarchical**, **DBSCAN**.
- Practice choosing **K** (elbow & silhouette) and reading **dendrograms**.
- Recognize common pitfalls and when to use each method.

# What Is Cluster Analysis?

- **Goal:** group *rows/subjects/cases* into unlabeled clusters based on similarity.
- **Outputs:** a categorical label (cluster ID) and optionally profiles/centroids.
- **Contrast with PCA/FA:** reduces **columns/variables** rather than rows.

## PCA/FA vs. Clustering (Big Picture)

- PCA/FA answers: *“Which variables hang together?”* → produces **components/factors** (new variables).
- Clustering answers: *“Which cases hang together?”* → produces **groups of cases** (new labels).
- Unit of analysis: PCA/FA → variables; Clustering → observations.

# Data as Coordinates

- Each case is a point in *p*-dimensional space.
- Similarity is operationalized via a **distance** in that space.
- Visual intuition: “clumps” of nearby points represent potential clusters.

```{r two-clumps-2d}
# Simulate two 2D clumps + a few specific cases
set.seed(123)
n <- 100
A <- cbind(rnorm(n,  0, 1), rnorm(n,  0, 1))
B <- cbind(rnorm(n,  4, 1), rnorm(n,  4, 1))
pts <- rbind(A, B)
plot(pts, pch = 19, main = "Two Clumps in 2D", xlab = "X1", ylab = "X2")
```

# Data Preparation

## Feature Scaling

- Always **scale** continuous features (mean 0, SD 1) before computing distances.
- Prevents a single large-scale variable from dominating.

```{r scale}
df <- scale(USArrests)  # built-in dataset: states × (Murder, Assault, UrbanPop, Rape)
head(df)
apply(df, 2, mean); apply(df, 2, sd)
```

## Distance Metrics

- **Euclidean** (L2): straight-line distance. Sensitive to **outliers**.
- **Manhattan** (L1): city-block distance. More **robust** to outliers; can yield square-ish clusters.
- Others: cosine, correlation distance, Gower (for mixed types).

```{r distances}
patient_demo <- data.frame(
  anxiety    = c(1, 4, 1, 3),
  depression = c(2, 6, 3, 5)
)
row.names(patient_demo) <- c("A","B","C","D")
patient_demo_scaled <- scale(patient_demo)

dist_euclid   <- dist(patient_demo_scaled, method = "euclidean")
dist_manhattan<- dist(patient_demo_scaled, method = "manhattan")
dist_euclid; dist_manhattan
```

# Method 1: K-Means Clustering

- Assumes roughly **spherical**, similarly sized clusters in scaled space.
- Minimizes within-cluster variance; initialization matters (use `nstart`).
- Pros: fast, simple; Cons: need **K** in advance; sensitive to outliers.

```{r kmeans-kselect, eval=FALSE}
library(factoextra)
set.seed(1)
df <- scale(USArrests)

# Elbow method
fviz_nbclust(df, kmeans, method = "wss") + ggplot2::labs(subtitle = "Elbow Method")

# Silhouette method
fviz_nbclust(df, kmeans, method = "silhouette") + ggplot2::labs(subtitle = "Silhouette Method")
```

```{r kmeans-run, eval=FALSE}
set.seed(123)
km <- kmeans(df, centers = 4, nstart = 25)
km$size; km$centers
factoextra::fviz_cluster(km, data = df) + ggplot2::labs(title = "K-means (K = 4)")
```

# Visualizing Cluster Profiles (Example)

```{r kmeans-profiles, eval=FALSE}
library(dplyr); library(tidyr); library(ggplot2)
USArrests_kmeans <- as.data.frame(df)
USArrests_kmeans$state   <- row.names(USArrests)
USArrests_kmeans$cluster <- factor(km$cluster)

cluster_profiles <- USArrests_kmeans |>
  dplyr::group_by(cluster) |>
  dplyr::summarise(across(Murder:Rape, mean)) |>
  tidyr::pivot_longer(Murder:Rape, names_to="Variable", values_to="Z") 

ggplot(cluster_profiles, aes(Variable, Z, group = cluster)) +
  stat_summary(fun = mean, geom = "line") +
  ggplot2::facet_wrap(~cluster) +
  ggplot2::labs(title = "Cluster Profiles (Z-scores)")
```

# Mapping Clusters (States Example)

```{r map, eval=FALSE}
library(ggplot2); library(dplyr); library(maps); library(mapproj)
cluster_data <- data.frame(
  region  = tolower(row.names(USArrests)),
  cluster = factor(km$cluster)
)
us_map <- map_data("state")
map_with_clusters <- left_join(us_map, cluster_data, by = "region")

ggplot(map_with_clusters, aes(x = long, y = lat, group = group, fill = cluster)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  coord_map("albers", lat0 = 39, lat1 = 45) +
  ggplot2::labs(title = "K-Means Clustering of US States (K = 4)", fill = "Cluster")
```

# Method 2: Hierarchical Clustering

- Build a tree of merges; no need to pick **K** up front.
- Key choices: **linkage** (e.g., Ward.D2, complete, average) and **distance**.
- Pros: dendrogram insight; Cons: greedy, harder to scale to very large *n*.

```{r hclust, eval=FALSE}
d  <- dist(df, method = "euclidean")
hc <- hclust(d, method = "ward.D2")
plot(hc, cex = 0.6, hang = -1, main = "Hierarchical Clustering — Ward.D2")
rect.hclust(hc, k = 4, border = 2:5)
clusters_hc <- cutree(hc, k = 4)
```

## Reading Dendrograms

- **Height** reflects dissimilarity at merge.
- **Long vertical jumps** suggest natural cut points.
- **Cut** to choose a desired number of clusters (e.g., `cutree(hc, k = 4)`).

# Method 3: DBSCAN (Density-Based)

- Defines clusters as dense regions separated by sparse regions.
- Parameters:
  - `eps`: neighborhood radius,
  - `minPts`: minimum points to form a dense core (often ≈ *p*+1 or 2*p*).
- Pros: finds **arbitrary shapes**, flags **outliers**; Cons: parameter tuning.

```{r dbscan, eval=FALSE}
library(dbscan)

# Heuristic: use k-NN distance plot to choose eps
k <- 4
kNNdistplot(df, k = k) + abline(h = 1.2, lty = 2)  # set a tentative eps

set.seed(1)
db <- dbscan::dbscan(df, eps = 1.2, minPts = k + 1)
table(db$cluster)   # 0 denotes noise
factoextra::fviz_cluster(list(data = df, cluster = db$cluster),
                         stand = FALSE, geom = "point",
                         ellipse = FALSE) + ggplot2::labs(title = "DBSCAN")
```

# Choosing Among Methods

- **K-means:** compact, spherical clusters; large *n*; need K.
- **Hierarchical:** exploratory structure; small–medium *n*; dendrogram needed.
- **DBSCAN:** varying density, noise/outliers, non-spherical shapes.

# Common Pitfalls & Tips

- Forgetting to **scale** features → distorted distances.
- Using Euclidean with heavy **outliers** → prefer Manhattan/robust methods.
- Blindly trusting elbow plots; corroborate with **silhouette** and domain sense.
- Evaluating clusters: **cohesion/separation**, profile interpretability, stability.
- Keep a **reproducible** pipeline: set seeds, record preprocessing choices.

# References / Further Reading

- Built-in example adapted from `USArrests` dataset and standard clustering workflows in R.
- See `?kmeans`, `?hclust`, `dbscan` package docs, and `factoextra` for visualization.
