---
title: "Cluster Analysis Project"
author: "YOUR_NAME"
date: "Data Science for the Psychological Sciences"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    css: AssignmentStyle.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.8,
  dpi = 150
)
library(tidyverse)
library(factoextra) 
library(dbscan)
```

# Project Overview

## Purpose

::: instructor-note
The purpose of this assignment is to apply **Cluster Analysis** to a psychological or health-related dataset. You will practice clustering data and then modeling those categorical outcomes (e.g., Yes/No, Pass/Fail, Diagnosed/Healthy) and interpreting results using log-odds and odds ratios.
:::

## Dataset Requirement

::: instructor-note
You must choose **one publicly available dataset**. Your dataset must include:

- At least **one binary outcome variable** (coded as 0 and 1).
- At least **two continuous predictors**.
- At least **one categorical predictor**.
:::

# Load and Inspect Data

```{r}
# Load your dataset here
# data <- read.csv("your_data.csv")
```

::: student-response
Describe your dataset in 2-4 sentences. What is the binary outcome you are predicting (the '1' vs the '0')?
:::

# Analysis 1: K-Means Clustering

## Research Question
::: student-response
Replace this text with your research question (e.g., "What are the natural groupings or subtypes of participants based on their scores across these variables?").
:::

## Data Preparation (Scaling)
::: instructor-note
Cluster analysis is highly sensitive to the scale of your variables. You must standardize your data before running any algorithms.
:::

```{r}
# Remove NAs, isolate your numeric variables, and use scale() to standardize the data
```

## Determining 'K'
```{r}
# Use fviz_nbclust() to create an Elbow plot or Silhouette plot to determine the optimal number of clusters
```

## Analysis & Visualization
```{r}
# 1. Run your k-means model here: km.res <- kmeans(df, centers = K, nstart = 25)
# 2. Visualize the clusters: fviz_cluster(km.res, data = df, geom = "point")
```

## Results & Interpretation
::: student-response
Report the result: 
1. What was your chosen value for K, and why did you choose it based on your heuristic plots? 
2. Look at the PCA cluster plot. Are the clusters well-separated, or do they heavily overlap?
3. Interpret one of your clusters: What defines the participants in this specific group?
:::

# Analysis 2: Hierarchical Clustering

## Research Question
::: student-response
Replace this text with your research question regarding the hierarchical "family tree" structure of your data.
:::

## Analysis 
```{r}
# 1. Calculate the distance matrix: d <- dist(df, method = "euclidean")
# 2. Run hierarchical clustering: hc.res <- hclust(d, method = "ward.D2")
```

## The Dendrogram
::: instructor-note
Use `fviz_dend()` to plot your hierarchical tree. Cut the tree into the same number of clusters (K) you used in Analysis 1 so you can compare the results.
:::

```{r}
# Create your dendrogram plot here
```

## Results
::: student-response
Look at the height (Y-axis) of the major merges on your dendrogram. Are there distinct "families" of data that merge very high up, indicating major differences between the groups? 
:::

# Analysis 3: Density-Based Clustering (DBSCAN)

## Research Question
::: student-response
How does accounting for density and isolating "noise" (outliers) change the structure of our clusters?
:::

## Analysis & Visualization
```{r}
# 1. Run DBSCAN here: db.res <- dbscan(df, eps = 1.25, minPts = 4)  *(Adjust parameters as needed)*
# 2. Visualize the results using fviz_cluster()
```

## Results & Reflection
::: student-response
1. Did DBSCAN identify any points as "noise" (Cluster 0)? 
2. Compare this plot to your K-Means plot. Because K-Means forces every single point into a cluster, it can sometimes distort the groups. Do the DBSCAN clusters look more tightly packed and "honest" now that the outliers have been removed?
3. Reflect on the data: Why might it make theoretical sense that some participants/cases in this dataset do not belong to *any* group?
:::

# Analysis 4: Evaluating the Clusters (External Validation)

## Research Question
::: student-response
Replace this text with your research question. (e.g., "Do the distinct cluster profiles we discovered differ significantly on Outcome Variable Y?")
:::

## Adding Cluster Labels to the Data
::: instructor-note
Before you can run a statistical test, you need to extract the cluster assignments from one of your models (e.g., your K-Means model) and attach them to your original dataframe as a new categorical independent variable.
:::

```{r}
# Extract the cluster assignments and add them as a new column
# Example: df$ClusterProfile <- as.factor(km.res$cluster)
```

## Analysis
::: instructor-note
Now that every participant is assigned to a "ClusterProfile", you can test if these profiles actually matter in predicting other variables!
- Use an **Independent t-test** if you have exactly 2 clusters and a continuous outcome.
- Use an **ANOVA** if you have 3 or more clusters and a continuous outcome.
- Use **Logistic Regression** if your outcome variable is binary (0/1).
:::

```{r}
# Run your chosen statistical test here (t.test, aov, or glm)
# Example: summary(aov(OutcomeVariable ~ ClusterProfile, data = df))
```

## Results & Reflection
::: student-response
1. Which statistical test did you choose to use based on your data, and why?
2. Was there a statistically significant difference between your clusters on the outcome variable? (Report your p-value).
3. Interpret the result: What does this tell you about the real-world validity or usefulness of the profiles you discovered?
:::