---
title: "Cluster Analysis — Lecture Summary"
subtitle: "From Data to Discovery: Data Science for the Psychological Sciences"
author: "Your Name"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  slidy_presentation:
    incremental: false
    self_contained: true
bibliography: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE, eval = FALSE, fig.width = 7, fig.height = 4.5)
```

```{r styles, results='asis', echo=FALSE}
cat('
<style>
body {
  font-family: "Helvetica Neue", Arial, sans-serif;
}

h1.title {
  color: #173F5F;
}

h2, h3 {
  color: #173F5F;
}

.slide {
  background: #FAFCFE;
}

.inverse {
  background: linear-gradient(135deg, #173F5F 0%, #20639B 100%);
  color: white;
}

.inverse h1, .inverse h2, .inverse h3, .inverse p, .inverse li {
  color: white !important;
}

.section-tag {
  display: inline-block;
  padding: 6px 12px;
  border-radius: 999px;
  background: #EAF2F8;
  color: #20639B;
  font-size: 18px;
  font-weight: 700;
  letter-spacing: 0.03em;
  margin-bottom: 12px;
}

.note-box {
  background: #FDF2E9;
  border-left: 8px solid #F39C12;
  padding: 14px 18px;
  border-radius: 8px;
  margin-top: 14px;
}

.takeaway-box {
  background: #EBF5FB;
  border-left: 8px solid #2E86C1;
  padding: 14px 18px;
  border-radius: 8px;
  margin-top: 14px;
}

.small {
  font-size: 0.78em;
}

.center {
  text-align: center;
}

.two-col {
  display: grid;
  grid-template-columns: 1fr 1fr;
  gap: 28px;
  align-items: start;
}

.left-col {
  font-size: 0.95em;
}

.right-col {
  text-align: center;
}

.slide ul, .slide ol {
  margin-top: 0.5em;
  margin-bottom: 0.5em;
  padding-left: 1.1em;
}

.slide li {
  margin-top: 0.3em;
  margin-bottom: 0.3em;
  line-height: 1.1;
}

/* Add a new one for wider text columns */
.two-col-wide-left {
  display: grid;
  grid-template-columns: 70% 30%;
  gap: 28px;
  align-items: start;
}

.three-col {
  display: grid;
  grid-template-columns: 1fr 1fr 1fr;
  gap: 20px;
  align-items: start;
  margin-top: 40px;
}
</style>
')
```

# Introduction to Cluster Analysis {.inverse}

::: {.center style="margin-top: 100px;"}
<h1 style="font-size: 3em;">From Data to Discovery</h1>
### Methods & Computing for the Psychological Sciences
<br>
**Topic:** Cluster Analysis
:::

# What is Cluster Analysis?

<div class="section-tag">Unsupervised Machine Learning</div>

Unlike supervised learning, where you have a known target variable (like price or species) to predict, clustering operates in a world without pre-existing labels.

- **The Primary Goal:** To discover natural groupings or clusters within your data.
- **The Application:** Identifying profiles or subtypes within a sample (e.g., grouping 500 patients into smaller sets of underlying groups like 'Mildly Anxious' or 'Healthy').

<div class="note-box">
**Common Applications:** Customer Segmentation, Anomaly Detection, Genomics, Identifying Psychiatric Subtypes, and Creating Cognitive/Personality Profiles.
</div>

# PCA vs. Cluster Analysis

<div class="section-tag">Two Types of Data Reduction</div>

While both are "data reduction" techniques, they reduce the data in fundamentally opposite ways:

::: {.two-col}

::: {.left-col}
### PCA / Factor Analysis
**Reduces Columns**

- **Answers:** "Which of my *variables* hang together?"
- **Goal:** Take many variables and find a smaller set of underlying components.
- **Analogy:** Sorting a bookshelf of 50 books by genre.
:::

::: {.right-col}
### Cluster Analysis
**Reduces Rows**

- **Answers:** "Which of my *subjects* hang together?"
- **Goal:** Take many subjects and find a smaller set of underlying profiles.
- **Analogy:** Sorting the 50 readers into groups based on the genres they read.
:::

:::

# The Core Principle

<div class="section-tag">The Golden Rule of Clustering</div>

<div class="center" style="margin-top: 60px;">
<h2 style="font-size: 2em; color: #173F5F;">
1. Participants <strong>within</strong> a cluster should be highly similar to one another.
<br><br>
2. Participants <strong>between</strong> different clusters should be highly dissimilar from one another.
</h2>
</div>

# Crucial Step: Data Preparation

<div class="section-tag">Feature Scaling</div>

Before you run an algorithm, you **must** prepare your data. Clustering distance metrics are highly sensitive to the scale of your variables.

- **The Problem:** Variables with large ranges (e.g., populations in the millions) will mathematically dominate variables with small ranges (e.g., percentages from 1 to 100).
- **The Solution:** You must scale your data so all variables contribute equally.
- **Z-score Standardization:** The standard approach in R using `scale()`.

```{r data-prep, echo=TRUE, message=FALSE, warning=FALSE}
# Load the built-in USArrests dataset and scale it
df <- scale(USArrests)
head(df)
```

# Three Main Families of Algorithms {.inverse}

::: {.center style="margin-top: 80px;"}

### 1. Partitioning Clustering (e.g., K-Means)
Partitions data into a pre-specified number of clusters (K). 

### 2. Hierarchical Clustering
Builds a tree-like hierarchy (dendrogram). Does not require you to specify K beforehand.

### 3. Density-Based Clustering (e.g., DBSCAN)
Groups closely packed points and identifies isolated points as "noise." Great for finding true outliers.

:::

# Partitioning: K-Means Clustering

<div class="section-tag">Minimizing Variance</div>

The goal of K-Means is to define clusters so that the total within-cluster variation (WCSS) is minimized. 

**How it works (Iterative Process):**
1. You specify the number of clusters (K).
2. The algorithm randomly drops K "centroids" into the data space.
3. **Assignment Step:** Every data point is assigned to its closest centroid.
4. **Update Step:** The centroid moves to the exact mathematical center (mean) of all the points assigned to it.
5. Repeat steps 3 & 4 until the centroids stop moving.

# Finding the Optimal 'K'

<div class="section-tag">Heuristics</div>

How do you pick the "right" number for K? We use heuristics to find the optimal balance.

::: {.two-col}

::: {.left-col}
### The Elbow Method
Plots the within-cluster sum of squares (WSS) against values of K. Look for the "elbow"—the point where adding another cluster doesn't significantly reduce error.

```{r elbow-plot, echo=FALSE, fig.height=3.5, message=FALSE, warning=FALSE}
library(factoextra)
fviz_nbclust(df, kmeans, method = "wss")
```
:::

::: {.right-col}
### The Silhouette Score
Calculates how cohesive a point is to its own cluster vs. how separated it is from the nearest neighboring cluster. Choose the K that maximizes this score.

```{r sil-plot, echo=FALSE, fig.height=3.5, message=FALSE, warning=FALSE}
fviz_nbclust(df, kmeans, method = "silhouette")
```
:::

:::

# Visualizing K-Means

<div class="section-tag">Dimensionality Reduction via PCA</div>

If our K-Means model uses 4 variables (Murder, Assault, Rape, UrbanPop), our clusters exist in 4D space. `fviz_cluster()` uses Principal Component Analysis (PCA) to create 2D axes that give us the "best angle" of cluster separation.

```{r kmeans-plot, echo=TRUE, fig.height=4.5, fig.width=8, fig.align='center', message=FALSE, warning=FALSE}
set.seed(123)
km.res <- kmeans(df, centers = 4, nstart = 25)

fviz_cluster(km.res, data = df,
             ellipse.type = "convex",
             geom = "point",
             ggtheme = theme_bw())
```

# Hierarchical Clustering

<div class="section-tag">Bottom-Up Agglomerative</div>

Hierarchical clustering builds a "family tree" without needing a pre-specified K.

**The Logic:**
1. Start: Treat every single data point as its own tiny cluster.
2. Find Closest: Find the two "closest" clusters in the dataset.
3. Merge: Merge those two into one new, larger cluster.
4. Repeat until all points are merged into one giant cluster.

**Ward's Method:** The most robust linkage method. Instead of just looking at physical distance, it merges the two clusters that result in the *minimum increase* in total within-cluster variance.

# Interpreting the Dendrogram

<div class="section-tag">The Family Tree</div>

::: {.two-col}

::: {.left-col}
- **X-axis (Leaves):** Individual data points (States).
- **Y-axis (Height):** The dissimilarity at which two clusters were merged. 
- **Reading the Tree:** Short U-shapes at the bottom indicate highly similar points merging. Long vertical lines near the top indicate major, distinct "families" of data.
:::

::: {.right-col}

```{r dendrogram, echo=TRUE, fig.height=5.5, message=FALSE, warning=FALSE}
d <- dist(df, method = "euclidean")
hc.res <- hclust(d, method = "ward.D2")

fviz_dend(hc.res, cex = 0.5, k = 4, 
          rect = TRUE, 
          main = "Hierarchical Clustering Dendrogram")
```

:::

:::

# Density-Based Clustering (DBSCAN)

<div class="section-tag">Honest Outliers</div>

DBSCAN defines clusters as crowded areas separated by empty space. 

**Why use DBSCAN?** K-Means and Hierarchical methods force *every single state* into a group, which can distort the data. DBSCAN identifies and isolates true outliers (like Alaska or Hawaii) as "noise", providing a more accurate reflection of reality.

```{r dbscan-plot, echo=TRUE, fig.height=4, fig.width=8, fig.align='center', message=FALSE, warning=FALSE}
library(dbscan)
set.seed(123)
# Note: eps and minPts must be defined based on your kNNdistplot in the chapter
db.res <- dbscan(df, eps = 1.25, minPts = 4) 

# Visualize, note how '0' (black points) are the outliers
fviz_cluster(db.res, data = df,
             ellipse = TRUE,
             geom = "point",
             ggtheme = theme_bw())
```

<div class="takeaway-box">
**Final Interpretation:** This might be the most "honest" result. K-Means and Hierarchical forced every state into a group. DBSCAN tells us that states like Florida, Alaska, California, and Nevada (which are often in the "noise" cluster) might be too unique to be grouped, leaving us with more homogenous, tightly-packed clusters.
</div>