---
title: "Answer Key: Exam — Introduction to R & RStudio"
author: ""
date: ""
output:
  pdf_document:
    toc: false
    number_sections: false
---

> This answer key corresponds to **intro_r_rstudio_exam_20_questions.Rmd**.

# Reference objects

```{r reference_objects, eval=FALSE}
x <- 10
y <- 5

ages <- c(25, 31, 19, 42, 50, 22)

student_data <- data.frame(
  ID = c(101, 102, 103, 104),
  Name = c("Alice", "Bob", "Charlie", "David"),
  Score = c(88, 92, 75, 85),
  Major = c("Biology", "History", "Math", "Biology")
)
```

# Part A — Describe the output

## 1) Output

- `5 + 3` prints **8**
- `10 - 4` prints **6**
- `6 * 7` prints **42**
- `20 / 4` prints **5**

---

## 2) Output

- `10 %% 3` prints **1**
- `%%` computes the **remainder** after division (modulus operator).

---

## 3) Output

With `x <- 10`, `y <- 5`:

- `x == 10` prints **TRUE**
- `y != 5` prints **FALSE**
- `x > y` prints **TRUE**

---

## 4) Output

- `(x > 5) & (y < 10)` prints **TRUE**
- `(x == 5) | (y == 5)` prints **TRUE**
- `!(x > y)` prints **FALSE**

---

## 5) Output

With `ages <- c(25, 31, 19, 42, 50, 22)`:

- `ages[3]` returns **19**
- `ages[c(1, 5)]` returns **c(25, 50)**
- `ages[2:4]` returns **c(31, 19, 42)**

---

## 6) Output

- `ages[seq(1, length(ages), by = 2)]` returns the odd-position elements: **c(25, 19, 50)**

---

## 7) Output

- `student_data$Score` returns **c(88, 92, 75, 85)**
- `mean(student_data$Score)` returns **85**

---

## 8) Output

- `student_data[3, ]` returns the **third row** (a 1-row data frame):  
  `ID = 103`, `Name = "Charlie"`, `Score = 75`, `Major = "Math"`
- `student_data[, 2]` returns the **Name column** (a character vector):  
  `c("Alice", "Bob", "Charlie", "David")`
- `student_data[3, 2]` returns the single value **"Charlie"**
- `student_data[, c("Name", "Major")]` returns a **data frame** with two columns (`Name`, `Major`) and all rows.

# Part B — Find and fix the error

## 9) Debugging

Correct comparison:

```{r a9, eval=FALSE}
x == 10
```

---

## 10) Debugging

Missing comma. Correct:

```{r a10, eval=FALSE}
student_data[3, ]
```

---

## 11) Debugging

Missing comma to indicate “all columns.” Correct:

```{r a11, eval=FALSE}
high_scorers <- student_data[student_data$Score > 85, ]
```

---

## 12) Debugging

Package name must be a character string:

```{r a12, eval=FALSE}
install.packages("dplyr")
```

---

## 13) Debugging

Missing closing parenthesis:

```{r a13, eval=FALSE}
library(dplyr)
```

---

## 14) Debugging

Search term must be a character string:

```{r a14, eval=FALSE}
help.search("correlation")
```

# Part C — Write the R code

## 15) Write code

```{r a15, eval=FALSE}
age <- 52
```

---

## 16) Write code

```{r a16, eval=FALSE}
name <- "Jordan"
```

---

## 17) Write code

```{r a17, eval=FALSE}
is_complete <- TRUE
```

---

## 18) Write code

```{r a18, eval=FALSE}
student_data[, c("Name", "Major")]
```

---

## 19) Write code

```{r a19, eval=FALSE}
biology_majors <- student_data[student_data$Major == "Biology", ]
```

---

## 20) Write code

```{r a20, eval=FALSE}
class(student_data)
str(student_data)
```
