---
title: "Quiz: Introduction to R & RStudio"
author: ""
date: ""
output:
  pdf_document:
    toc: false
    number_sections: false
---



> NAME: _____________________________________

# Reference objects (for questions that use them)

Copy/paste the following into R **if you want to test your thinking**, but the exam questions can be answered without running code.

```{r reference_objects, eval=FALSE}
# Variables for operator questions
x <- 10
y <- 5

# A vector for indexing/slicing questions
ages <- c(25, 31, 19, 42, 50, 22)

# A simple data frame for data-frame questions
student_data <- data.frame(
  ID = c(101, 102, 103, 104),
  Name = c("Alice", "Bob", "Charlie", "David"),
  Score = c(88, 92, 75, 85),
  Major = c("Biology", "History", "Math", "Biology")
)
```

## Question 1 (Write code)

Create a character variable named `name` that stores the text `"Jordan"`.

```{r q16, eval=FALSE}
# YOUR CODE


```

---

## Question 2 (Output)

What printed output do you expect from each line?

```{r q1, eval=FALSE}
5 + 3

10 - 4

6 * 7

20 / 4

2 ** 2

10 %% 10
```

---

\newpage

## Question 3 (Output)

For each line, describe what the command is doing and what the output would be (if any)
```{r q3, eval=FALSE}
x <- 10

y <= 5

x > 10

y == 10

x > y
```

---

## Question 4 (Output)

What printed output do you expect from each line?
```{r q4, eval=FALSE}
(x > 5) & (y < 10)

(x == 5) | (y == 5)

!(x > y)
```

---

## Question 5 (Output)

What printed output do you expect from each line?
```{r q5, eval=FALSE}
ages[3]

ages[c(1, 5)]

ages[2:4]
```

---

## Question 6 (Output)
What printed output do you expect from each line?

```{r q7, eval=FALSE}
student_data$Score[2]

student_data[1,2]
```


---

## Question 7 (Debugging)

A student tries to subset (i.e., extract) the third **row** of a data frame but gets an error.

Identify the mistake and provide a corrected version that returns the **third row**.

```{r q10, eval=FALSE}
student_data[3 ]





```

---

## Question 8 (Debugging)

A student tries to install a package:

What is the error? Provide the corrected command.

```{r q12, eval=FALSE}
install.packages(dplyr)



```

---

## Question 9 (Debugging)

A student tries to search the help system for a function to compute a correlation:

What is the issue? Provide the corrected command.
```{r q14, eval=FALSE}
help.search(correlation)



```

---

## Question 10 (Write code)

Using `student_data`, write a single line that returns only the `Name` and `Major` columns (all rows).

```{r q18, eval=FALSE}
# YOUR CODE



```

---

## Question 11 (Write code)

Using `student_data`, write code that creates a **new object** named `biology_majors` containing only rows where `Major` is `"Biology"`.

```{r q19, eval=FALSE}
# YOUR CODE


#-
```

---