---
title: "Exam: Introduction to R & RStudio"
author: ""
date: ""
output:
  pdf_document:
    toc: false
    number_sections: false
---

> Instructions: Answer all questions. Unless a question explicitly asks for narrative, you may respond with a brief sentence or a short R command/code snippet.

# Reference objects (for questions that use them)

Copy/paste the following into R **if you want to test your thinking**, but the exam questions can be answered without running code.

```{r reference_objects, eval=FALSE}
# Variables for operator questions
x <- 10
y <- 5

# A vector for indexing/slicing questions
ages <- c(25, 31, 19, 42, 50, 22)

# A simple data frame for data-frame questions
student_data <- data.frame(
  ID = c(101, 102, 103, 104),
  Name = c("Alice", "Bob", "Charlie", "David"),
  Score = c(88, 92, 75, 85),
  Major = c("Biology", "History", "Math", "Biology")
)
```

\newpage

# Part A — Describe the output

## Question 1 (Output)

```{r q1, eval=FALSE}
5 + 3
10 - 4
6 * 7
20 / 4
```

What printed output do you expect from each line?

---

## Question 2 (Output)

```{r q2, eval=FALSE}
10 %% 3
```

What is printed, and what does the operator `%%` compute?

---

## Question 3 (Output)

```{r q3, eval=FALSE}
x <- 10
y <- 5
x == 10
y != 5
x > y
```

List the printed TRUE/FALSE outputs in order.

---

## Question 4 (Output)

```{r q4, eval=FALSE}
(x > 5) & (y < 10)
(x == 5) | (y == 5)
!(x > y)
```

List the printed TRUE/FALSE outputs in order.

---

## Question 5 (Output)

```{r q5, eval=FALSE}
ages[3]
ages[c(1, 5)]
ages[2:4]
```

What does each line return?

---

## Question 6 (Output)

```{r q6, eval=FALSE}
ages[seq(1, length(ages), by = 2)]
```

What does this return? (Write the vector values.)

---

## Question 7 (Output)

```{r q7, eval=FALSE}
student_data$Score
mean(student_data$Score)
```

What does each line return?

---

## Question 8 (Output)

```{r q8, eval=FALSE}
student_data[3, ]
student_data[, 2]
student_data[3, 2]
student_data[, c("Name", "Major")]
```

Briefly describe what each line returns (e.g., “a data frame row”, “a character vector”, “a single value”, etc.) and the specific values you expect.

\newpage

# Part B — Find and fix the error

## Question 9 (Debugging)

A student tries to compare two values:

```{r q9, eval=FALSE}
x <- 10
x = 10
```

This does not perform a comparison. What is the correct line of code to test whether `x` is equal to 10?

---

## Question 10 (Debugging)

A student tries to subset a data frame but gets an error:

```{r q10, eval=FALSE}
student_data[3 ]
```

Identify the mistake and provide a corrected version that returns the **third row**.

---

## Question 11 (Debugging)

A student tries to filter to high scorers but the code fails:

```{r q11, eval=FALSE}
high_scorers <- student_data[student_data$Score > 85 ]
```

Identify the missing piece(s) and provide a corrected line that returns all rows with `Score > 85`.

---

## Question 12 (Debugging)

A student tries to install a package:

```{r q12, eval=FALSE}
install.packages(dplyr)
```

What is the error? Provide the corrected command.

---

## Question 13 (Debugging)

A student tries to load a package:

```{r q13, eval=FALSE}
library(dplyr
```

What is the error? Provide the corrected command.

---

## Question 14 (Debugging)

A student tries to search the help system:

```{r q14, eval=FALSE}
help.search(correlation)
```

What is the issue? Provide the corrected command.

\newpage

# Part C — Write the R code

## Question 15 (Write code)

Create a variable named `age` that stores the value 52 using the conventional R assignment operator.

```{r q15, eval=FALSE}
# YOUR CODE
```

---

## Question 16 (Write code)

Create a character variable named `name` that stores the text `"Jordan"`.

```{r q16, eval=FALSE}
# YOUR CODE
```

---

## Question 17 (Write code)

Create a logical variable named `is_complete` that stores the value `TRUE`.

```{r q17, eval=FALSE}
# YOUR CODE
```

---

## Question 18 (Write code)

Using `student_data`, write a single line that returns only the `Name` and `Major` columns (all rows).

```{r q18, eval=FALSE}
# YOUR CODE
```

---

## Question 19 (Write code)

Using `student_data`, write code that creates a new object named `biology_majors` containing only rows where `Major` is `"Biology"`.

```{r q19, eval=FALSE}
# YOUR CODE
```

---

## Question 20 (Write code)

Write one line of code to check the **class** of `student_data`, and a second line of code to view its structure.

```{r q20, eval=FALSE}
# YOUR CODE (2 lines)
```
