---
title: "Applied Linear Models Project"
author: "YOUR_NAME"
date: "Data Science for the Psychological Sciences"
output:
  html_document:
    toc: false
    toc_float: true
    number_sections: true
    css: AssignmentStyle.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.8,
  dpi = 150
)
```

# Project Overview

## Purpose

::: instructor-note
The purpose of this assignment is to apply the general linear model to a dataset of your choice. You will identify research questions that can be answered using correlation/regression, a t-test, and ANOVA, then use R to wrangle, visualize, analyze, and interpret your data.
:::

## Dataset Requirement

::: instructor-note
You must choose **one publicly available dataset** and use it for the entire project.

Your dataset must include enough variables to support all four analyses below:

-   at least **two continuous variables**
-   at least **one categorical variable with two levels** for a t-test
-   at least **one categorical variable with three or more levels** for an ANOVA
:::

## Deliverable

::: instructor-note
Submit **one knitted HTML or PDF document** that includes:

-   all annotated R code
-   all output
-   all figures
-   written answers and interpretations for each section

*Your submission should be clear enough that another student could follow your workflow from start to finish.*
:::

## Required Analyses

::: instructor-note
You must complete **4 total analyses**:

1.  **Correlation or simple linear regression**
2.  **A t-test**
3.  **An ANOVA**
4.  **One additional analysis of your choice** from the following:
    -   another correlation/regression
    -   another t-test
    -   another ANOVA
:::

# Load Your Data

```{r}
# Load your dataset here
```

## Initial Inspection

Inspect the structure of your data.

```{r}
# Example:
# glimpse(your_data)
# or
# summary(your_data)
```

::: instructor-note
Write 2-4 sentences describing what kinds of variables are available and which ones you may use in your analyses.
:::

::: student-response
Replace this text with your hypothesis.
:::

# Analysis 1: Correlation or Regression

## Research Question

::: student-response
Replace this text with your research question/hypothesis in 1-2 sentences.
:::

## Variables

::: instructor-note
Identify:

-   your predictor variable
-   your outcome variable
-   whether each variable is continuous or categorical
:::

::: student-response
Replace this text with the description of your variables
:::

## Data Wrangling

::: instructor-note
Create a clean dataset for this analysis.

Your wrangling should:

-   keep only the variables needed for this analysis
-   remove missing values as needed
-   rename variables if helpful
-   convert variable types if necessary
:::

```{r}
# Wrangle your data for Analysis 1 here
```

Briefly explain what you changed and why.

::: student-response
Replace this text with your explanation.
:::

## Visualization

::: instructor-note
Create at least:

-   one plot showing the relationship between the variables
-   one plot showing the distribution of the variables
-   or use gg_extra to show both in a single plot

Examples may include:

-   scatterplot
-   histogram
-   density plot
-   boxplot
:::

```{r}
# Create your visualizations here
```

::: student-response
Describe what you notice in 2-4 sentences.
:::

## Analysis

::: instructor-note
Conduct either:

-   a correlation analysis, or
-   a simple linear regression
:::

```{r}
# Run your analysis here
```

## Results

::: instructor-note
Report the main result in words.

Be sure to describe:

-   whether the relationship was statistically significant
-   the direction of the relationship
-   the size of the relationship, if relevant
-   what the result means in plain English
:::

::: student-response
Write 3-5 sentences.
:::

# Analysis 2: T-Test

## Research Question

::: student-response
Replace this text with your research question/hypothesis in 1-2 sentences.
-   What continuous outcome are you comparing?
-   What two groups are you comparing?
:::

## Variables

::: instructor-note
Identify:

-   your grouping variable
-   your outcome variable
-   whether this is a **between-subjects** or **within-subjects** comparison
:::

::: student-response
Replace this text with a description of your variables
:::

## Data Wrangling

::: instructor-note
Create a clean dataset for this analysis.

Your wrangling should:

-   keep only the variables needed
-   remove missing values
-   make sure the grouping variable is coded correctly
-   make sure there are exactly two groups being compared
:::

```{r}
# Wrangle your data for Analysis 2 here
```

Briefly explain what you changed and why.

::: student-response
Replace this text with your explanation.
:::

## Visualization

::: instructor-note
Create an appropriate plot for your t-test.

Examples include:

-   boxplot with jitter
-   violin plot
-   mean/error bar plot
:::

```{r}
# Create your visualization here
```

Describe what you notice in 2-4 sentences.

::: student-response
Replace this text with your description
:::

## Analysis

::: instructor-note
Conduct a t-test appropriate for your design.

Examples:

-   `t.test(..., var.equal = TRUE)` for an independent-samples t-test
-   `t.test(..., paired = TRUE)` for a paired-samples t-test
:::

```{r}
# Run your t-test here
```

## Results

::: instructor-note
Report the main result in words.

Be sure to describe:

-   whether the group difference was statistically significant
-   which group had the higher mean
-   what the result means in plain English
:::

::: student-response
Replace this text with the description of your results.
:::

# Analysis 3: ANOVA

## Research Question

::: student-response
Replace this text with your research question/hypothesis in 1-2 sentences.
:::

## Variables

::: instructor-note
Identify:

-   your grouping variable
-   your outcome variable
-   whether this is a **between-subjects one-way ANOVA** or a **within-subjects one-way ANOVA**
:::

::: student-response
Replace this text with the description of your variables
:::

## Data Wrangling

::: instructor-note
Create a clean dataset for this analysis.

Your wrangling should:

-   keep only the variables needed
-   remove missing values
-   make sure the grouping variable is a factor
-   make sure the factor has at least 3 levels
:::

```{r}
# Wrangle your data for Analysis 3 here
```

Briefly explain what you changed and why.

::: student-response
Replace this text with your explanation.
:::

## Visualization

::: instructor-note
Create an appropriate plot for your ANOVA.

Examples include:

-   boxplot with jitter
-   violin plot
-   line plot for repeated-measures data
:::

```{r}
# Create your visualization here
```

Describe what you notice in 2-4 sentences.

::: student-response
Replace this text with your description
:::

## Analysis

::: instructor-note
Conduct an ANOVA appropriate for your design.

Examples:

-   `aov()` for a one-way ANOVA
-   repeated-measures ANOVA if your design is within-subjects
:::

```{r}
# Run your ANOVA here
```

::: instructor-note
If your omnibus ANOVA is significant, conduct a follow-up comparison procedure.

Examples:

-   `TukeyHSD()`
-   pairwise comparisons with p-value adjustment
:::

```{r}
# Run post-hoc or follow-up tests here if needed
```

## Results

::: instructor-note
Report the main result in words.

Be sure to describe:

-   whether the overall ANOVA was statistically significant
-   whether there were differences among groups
-   which groups differed, if you conducted follow-up tests
-   what the result means in plain English
:::

::: student-response
Replace this text with the description of your results.
:::

# Analysis 4: Additional Analysis of Your Choice

::: instructor-note
Choose **one additional analysis** from the following:

-   correlation/regression
-   t-test
-   ANOVA

This analysis must be different from at least one of your earlier examples in either:

-   variables used
-   research question
-   design
:::

## Research Question

::: student-response
Replace this text with your research question/hypothesis in 1-2 sentences.
:::

## Variables

Identify the variables you will use and their types.

::: student-response
Replace this text with a description of your variables
:::

## Data Wrangling

```{r}
# Wrangle your data for Analysis 4 here
```

Briefly explain what you changed and why.

::: student-response
Replace this text with your brief explanation
:::

## Visualization

```{r}
# Create your visualization(s) here
```

Describe what you notice in 2-4 sentences.

::: student-response
Replace this text with your description
:::

## Analysis

```{r}
# Run your analysis here
```

## Results

Report the main result in words.

::: student-response
Replace this text with 3-5 sentences explaining the result in plain English.
:::
