---
title: "Applied Logistic Models Project"
author: "YOUR_NAME"
date: "Data Science for the Psychological Sciences"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    css: AssignmentStyle.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.8,
  dpi = 150
)
library(tidyverse)
library(broom) 
library(NHANES)
```

# Project Overview

## Purpose

::: instructor-note
The purpose of this assignment is to apply **Logistic Regression** to a psychological or health-related dataset. You will practice modeling binary outcomes (e.g., Yes/No, Pass/Fail, Diagnosed/Healthy) and interpreting results using log-odds and odds ratios.
:::

## Dataset Requirement

::: instructor-note
You must choose **one publicly available dataset**. Your dataset must include:

- At least **one binary outcome variable** (coded as 0 and 1).
- At least **two continuous predictors**.
- At least **one categorical predictor**.
:::

# Load and Inspect Data

```{r}
# Load your dataset here
# data <- read.csv("your_data.csv")
```

::: student-response
Describe your dataset in 2-4 sentences. What is the binary outcome you are predicting (the '1' vs the '0')?
:::

# Analysis 1: Simple Logistic (Continuous Predictor)

## Research Question
::: student-response
Replace this text with your research question (e.g., "Does variable X predict the probability of outcome Y?").
:::

## Data Wrangling
```{r}
# Ensure outcome is numeric 0/1 and remove NAs
```

## Visualization
::: instructor-note
Create a plot using `geom_jitter()` and `geom_smooth(method = "glm", method.args = list(family = "binomial"))`.
:::

```{r}
# Create your S-Curve plot here
```

## Analysis & Results
```{r}
# Run your glm() model here: model1 <- glm(outcome ~ var, data = df, family = "binomial")
```

::: student-response
Report the result: 
1. Is the predictor significant? 
2. What is the **Odds Ratio** (exp of the coefficient)? 
3. Interpret: "For every 1-unit increase in [X], the odds of [Y] change by [Z]%."
:::

# Analysis 2: Simple Logistic (Categorical Predictor)

## Research Question
::: student-response
Replace this text with your research question comparing two or more groups.
:::

## Analysis 
```{r}
# Run your categorical glm() model here
```

## Results
::: student-response
Report the Odds Ratio comparing your groups. Which group is more likely to experience the "1" outcome?
:::

# Analysis 3: Multiple Logistic Regression

## Research Question
::: student-response
How do your continuous and categorical predictors work together to predict the outcome?
:::

## Analysis
```{r}
# Run your multiple logistic regression here: glm(outcome ~ var1 + var2, family = "binomial")
```

## Classification Table (Confusion Matrix)
```{r}
# Use predict(..., type = "response") to create a confusion matrix at 0.50 threshold
```

## Results & Reflection
::: student-response
1. Were both predictors significant when included together?
2. Look at your Confusion Matrix. Did the model make more False Positives or False Negatives?
3. Reflect on the **Base Rate**: If your model was significant but the probabilities stayed low, why?
:::